# Clipping with gIntersects

An alternative to `over` for spatial subsetting is `gIntersects`,
although we could equally use 
`gContains`, `gWithin` and other `g...` functions. 
The power of these commands can be seen by accessing the 
rgeos help pages, e.g. `?gOverlaps`.
`gIntersects` will output information for each point, telling us which 
polygon it interacts with (i.e. the polygon it is in):

```{r Identifying and plotting individual stations, results='hide'}
int <- gIntersects(stations, lnd, byid = T) # find which stations intersect 
class(int) # it's outputed a matrix
dim(int) # with 33 rows (one for each zone) and 2532 cols (the points)
summary(int[,c(200,500)]) # not the output of this
plot(lnd)
points(stations[200,], col = "red") # note point id 200 is outside the zones
points(stations[500,], col = "green") # note point 500 is inside
which(int[,500] == T) # this tells us that point 500 intersects with zone 32
points(coordinates(lnd[32,]), col = "black") # test the previous statement
```

In the above code, only the first line actually 'does' anything
in our workspace, by creating the object `int`. The proceeding 
lines are dedicated to exploring this object and what it means. 
Note that it is a matrix with columns corresponding to the points and 
rows corresponding to boroughs. The borough in which a particular 
point can be extracted from `int` as we shall see below.
For the purposes of clipping, we are only interested in whether
the point intersects with _any_ of the boroughs. This is where the 
function `apply`, which is unique to R, comes into play:

```{r Clipped points (within London boroughs)}
clipped <- apply(int == F, MARGIN = 2, all)
plot(stations[which(clipped),]) # shows all stations we DO NOT want
stations.cl <- stations[which(!clipped),] # use ! to select the invers
points(stations.cl, col = "green") # check that it's worked
stations <- stations.cl; rm(stations.cl) # tidy up: we're only interested in clipped ones
```

The first line instructs R to look at each column (`MARGIN = 2`, we would use
`MARGIN = 1` for row-by-row analysis) and report back whether `all` of the values are
false. This creates the inverse selection that we want, hence the use of `!` to invert it.
We test that the function works on a new object (often a good idea, to avoid overwriting 
useful data) with plots and, once content that the clip has worked, save the sample of 
points to our main `stations` object and remove the now duplicated `stations.cl` object.
