---
title: "A practical introduction to R and RStudio"
author: "Robin Lovelace"
date: "`r Sys.Date()`"
output: ioslides_presentation
bibliography: creating-maps.bib
---

```{r, include=FALSE}
pkgs <- c("png", "grid", "shiny", "knitr")
lapply(pkgs, library, character.only = T)
# opts_knit$set(root.dir = "~/repos/Creating-maps-in-R/")
```

# An introduction to RStudio | Pages 1:5 of Introduction to visualising spatial data in R

## What is RStudio and and why use it?

What is RStudio?

- An IDE
- A project management system
- A document preparation system
- An online publication portal


## Installing and using R and RStudio

- Install the latest version of R: http://cran.rstudio.com/ 
- Install the latest version of RStudio for your system:
http://www.rstudio.com/products/rstudio/download/
- Get acquainted to RStudio and the panels - play around!
http://dss.princeton.edu/training/RStudio101.pdf

```{r, echo=FALSE}
grid.raster(readPNG("figure/rstudio-proj.png"))
```

## Productivity with RStudio 

Key shortcuts in RStudio:

```{r, echo=FALSE}
shortcuts <- data.frame(Command = c(
  "Alt + Shift + K", 
  "Ctrl + Enter",
  "Ctrl + R",
  "Tab"),
  Action = c("Show shortcuts",
    "Run current line of code",
    "Run all lines of code in the script",
    "Autocomplete*"))
kable(shortcuts)
```

## Code used to generate that table | a mini demo of RMarkdown

RStudio is not just about R - it's a productivity suite!

```{r, eval=FALSE}
shortcuts <- data.frame(Command = c(
  "Alt + Shift + K", 
  "Ctrl + Enter",
  "Ctrl + R",
  "Tab"),
  Action = c("Show shortcuts",
    "Run current line of code",
    "Run all lines of code in the script",
    "Autocomplete*"))
kable(shortcuts)
```

## R packages

There are 7,000+ 'add-on' packages to 'supercharge' R.

Easiest way to install them, from RStudio:

```
Tools -> Install Packages
```

or using keyboard shortcuts:

```
Alt + T ... then k
```

## Install packages for this tutorial

Can be installed and loaded in 6 lines of code:

```{r, eval=FALSE}
pkgs <- c("devtools", "shiny", "rgdal", "rgeos", "ggmap", "raster")
install.packages(pkgs) # install the official packages!
library(devtools) # enables installation of leaflet
gh_pkgs <- c("rstudio/leaflet", "robinlovelace/stplanr") 
install_github(gh_pkgs) # install packages on github
lapply(c(pkgs, "leaflet", "stplanr"), library, character.only = T)
```

```{r, include=FALSE}
pkgs <- c("devtools", "shiny", "rgdal", "rgeos", "ggmap", "raster")
library(devtools) # enables installation of leaflet
lapply(c(pkgs, "leaflet", "stplanr"), library, character.only = T)
```

## Features of RStudio

> - Flexible window pane layouts to optimise use of screen space and enable fast interactive visual feed-back.
> - Intelligent auto-completion of function names, packages and R objects.
> - A wide range of keyboard shortcuts.
> - Visual display of objects, including a searchable data display table.
> - Real-time code checking and error detection.
> - Menus to install and update packages.
> - Project management and integration with version control.

## RStudio panes

RStudio has four main window 'panes':

 - The **Source pane**, for editing, saving, and dispatching R code to the console (top left). 
 
 - The **Console pane**. Any code entered here is processed by R, line by line (bottom left).
 
 - The **Environment pane** (top right) contains information about the current objects loaded in the workspace including their class, dimension (if they are a data frame) and name. 
 
 - The **Files pane** (bottom right) contains a simple file browser, a Plots tab, Help and Package tabs and a Viewer.
 
### Exercises

You are developing a project to visualise data.
Test out the multi-panel RStudio workflow by following the steps below:

1. Create a new folder for the input data using the **Files pane**.

2. Type in `downl` in the **Source pane** and hit `Enter` to make the function `download.file()` autocomplete. Then type `"`, which will autocomplete to `""`.

3. Execute the full command with `Ctrl-Enter`:

    ```{r, eval=FALSE}
    download.file("https://www.census.gov/2010census/csv/pop_change.csv",
                  "data/pop_change.csv")
    ```

3. Write and execute a command to read-in the data, such as

    ```{r}
    pop_change = read.csv("https://www.census.gov/2010census/csv/pop_change.csv", skip = 2)
    ```

    ```{r, echo=FALSE}
    # question: complete example with plots and all here?
    ```

4. Use the **Environment pane** to click on the data object `pop_change`. 

5. Use the **Console** to test different plot commands to visualise the data, saving the code you want to keep back into the **Source pane**, as `pop_change.R`.

6. Use the **Plots tab** in the Files pane to scroll through past plots. Save the best using the Export dropdown button.

## Project management

In the far top-right of RStudio there is a diminutive drop-down menu illustrated with R inside a transparent box.

Projects

- Set the working directory automatically. `setwd()`, a common source of error for R users, is rarely if ever needed.

- The last previously open file is loaded into the Source pane. 

- The `File` tab displays the associated files and folders in the project, allowing you to quickly find your previous work.

- Any settings associated with the project, such as Git settings, are loaded. This assists with collaboration and project-specific set-up.

## Setting up a project for the remainder of this course

- Go to https://github.com/Robinlovelace/Creating-maps-in-R

# Objects, functions and concepts for efficient R programming 


## Basic data types

Anything that exists in R is an object. Let's create some with the `<-` symbol (`=` does the same job, before you ask!)

```{r}
vector_logical <- c(TRUE, TRUE, FALSE)
vector_character <- c("yes", "yes", "Hello!")
vector_numeric <- c(1, 3, 9.9)

class(vector_logical) # what are the other object classes?
```

Use the "Environment tab" (top right in RStudio) to see these

## Intermediate data types

R has a hierarchy of data classes, tending to the lowest:

- Binary
- Integer (numeric)
- Double (numeric)
- Character

## Examples of data types

```{r}
a <- TRUE
b <- 1:5
c <- pi
d <- "Hello Leeds"
```

```{r, eval=FALSE}
class(a)
class(b)
class(c)
class(d)
```

## Class coercion I

```{r}
ab <- c(a, b)
ab
class(ab)
```

## Class coercion II

- R automatically forces some objects into the class it thinks it is best
- Demo:

```{r}
x = 1:5
class(x)
x = c(x, 6.1)
class(x)
```

> - Test: what is the class of `x = c(x, "hello")`?

## Dimensions of objects

- Dimensionality is key to understanding R data
- You cannot do the same thing with a square as you can a line
- R is 'vectorised', meaning it deals with many numbers at once

Test: what is the dimension of objects we created in the last slide?

## Vectorised code I

Python is not vectorised by default, hence:

```{r, engine='python'}
a = [1,2,3]
b = [9,8,6]
print(a + b)
```

## Vectorised code II

R is vectorised, meaning that it adds each element automatically

```{r}
a = c(1,2,3)
b = c(9,8,6)
a + b
```

- The same applies to matrices - R understands matrix algebra
- See `?matmult` for more on matrix multiplication

## Vectorised code III

```{r}
x <- c(1, 2, 5)
for(i in x){
  print(i^2)
}
```

Creating a new vector based on `x`

```{r}
for(i in 1:length(x)){
  if(i == 1) x2 <- x[i]^2
  else x2 <- c(x2, x[i]^2)
}
x2
```

## Test on data types

```{r}
class(c(a, b))
class(c(a, c))
class(c(b, d))
```

## Sequences

```{r}
x <- 1:5
y <- 2:6
plot(x, y)
```

## Sequences with seq

```{r}
x <- seq(1,2, by = 0.2)
length(x)
x <- seq(1, 2, length.out = 5)
length(x)
```

## The data frame

The fundamental data object in R. 

Create them with `data.frame()`

```{r}
data.frame(vector_logical, vector_character, n = vector_numeric)
```

Oops - we forgot to assign that. Tap `UP` or `Ctl-UP` in the console, then enter:

```{r}
df <- data.frame(vector_logical, vector_character, n = vector_numeric)
```

## Common dimensions

|    | Homogeneous   | Heterogeneous |
|----|---------------|---------------|
| 1d | Atomic vector | List          |
| 2d | Matrix        | Data frame    |
| nd | Array         |               |

Source: @wickham_advanced_2014

## Automating things

To ask R what objects it has, we can use `ls()`.

(Anything that happens is a function)

```{r}
ls()
```

Now we can automate the question: what class?

```{r}
obs <- ls()[grep("ve", ls())]
sapply(X = mget(obs), FUN = class)
```

## Getting help in R

To find out what just happened, we can use R's internal help

The most commonly used help functions are:

```{r}
help(apply) # get help on apply
?apply 
?sapply
??apply
```

The `*apply` family of functions are R's internal `for` loops. What about `get()`

```{r}
?get
```

## Plotting data in R

`plot()` is polymorphic. Try `plot(df)` and `?plot`:

```{r, echo=FALSE}
plot(df) # note: now we can do things with the df
?plot    # get help on plotting
```

# Data manipulation and plotting paradigms

## Subsetting data in R

The `[]` brackets, appending the object name, subset data.

A comma separates each dimension; nothing means everything:

```{r}
df[1,] # all of the the 1st line of df
```

In a 2d dataset, the following selects the 3rd line in the 3rd column:

```{r}
df[3,3]
```

## Manipulating columns

New columns can be created as follows:

```{r}
df$new_col = NA
```

Or as a function of old ones:

```{r}
df$new_col = df$vector_logical + df$n
```

## Exercise: manipulate and plot the pop_change dataset


