---
title: "R Spatial Reference Card"
author: "Robin"
date: "10/07/2014"
output: pdf_document
---

R has powerful and effective spatial data processing tools. It is the purpose
of this reference card to summarise the most important and frequently used
packages and functions. Some of the commands use real datasets from the
'Creating-maps-in-R' GitHub repository. Unzip this repo and run the commands from
the folder's root directory to see the commands in action, or use R's help functions.
Using the example of the `help` command, type:
`?help` (for information on a specific function),
`??help` (to search for word in R's documentation)
and `example(help)` (for premade examples of that function).

# Key packages

Additional packages must be installed (using `install.packages('packageName')`) on top of R's [default packages](http://www.hep.by/gnu/r-patched/r-admin/R-admin_33.html) to use R's spatial capabilities. Some of the most important and frequently loaded, via `library(packageName)`,
spatial packages are:

- **sp**: basis of most other spatial R packages - provides Spatial* classes
- **rgdal**: R's interface to the GDAL data import library
- **rgeos**: provides a number of spatial functions
- **maptools**: tools for handling spatial objects
- **raster**: provides functions for raster data

To refer to a function in relation to its package, use the `::` symbol.
`sp::spTransform`, for example, is the same as `spTransform`, but avoids
any ambiguity by specifying the package of which it is a part. There are
hundreds of additional R packages described on the Comprehensive
R Archive Network's (CRAN) website under the
[*spatial view*](http://cran.r-project.org/web/views/Spatial.html).

# R's Spatial* classes

To work with spatial data, R uses *classes* that are more complex
that the default *S3* `vector`, `list` and `data.frame` data classes
in R's base packages. Spatial object classes can be identified with the
function `class(objectName)`. Each contain a number of different
*slots*: sub-classes of specific types. Some important
spatial classes provided by the `sp` package include:

- `SpatialPoints`: point data, with x and y coordinates for each item
- `SpatialLines`: line data, composed of lists of the `Lines` sub-class or *slot*
- `SpatialPolygons`: polygon data, with `Polygons` and `Polygon` *slots*.
- `SpatialPixels`: a spatial class for raster data

It is important to note that each of the above classes cannot contain non-spatial
attribute data. First they must be converted into a class of the same name but suffixed
with `DataFrame`. Thus `spPdf <- SpatialPointsDataFrame(spP, df)` will, provided
`nrow(spP) == nrow(df)` and that `spP` and `df` are of `SpatialPoints` and `data.frame`
classes, respectively, create a new object containing spatial and non-spatial information.
The attribute data can be accessed in this instance by `spPdf@data`: the `@data` notation
refers to the `data` slot of the new object. More fundamental spatial classes, which
are in fact subsets of the `Spatial*` type classes listed above, include `bbox` (the
object's bounding box) and `proj4string` (the projection of the object, which can be `NA`).

# Loading and querying spatial data

```{r, warning=FALSE, message=FALSE, results='hide'}
library(rgdal)
SpDf <- readOGR("data/", "london_sport") # this loads a shapefile
```

# Allocating and changing projection

Query the current projection of an object: 
```{r}
proj4string(SpDf)
```

# Spatial subsetting

# Spatial aggregation

# Spatial graphics

# Basemaps and advanced functions


