---
title: "R Spatial Reference Card"
author: "Robin Lovelace and Rachel Oldroyd"
output: pdf_document
geometry: margin=0.3in
---

```{r global_options, include=FALSE}
library(knitr)
opts_chunk$set(fig.width=12, fig.height=8, fig.path='Figs/',
warning=FALSE, message=FALSE, results='hide')
# remove or change these options to see the output of the commands
```

R has a suite of powerful and effective spatial data processing tools. This reference card to summarises the fundamentals of R for geographical applications, in terms of packages and functions. The code has been tested on data from the
'Creating-maps-in-R' [GitHub repository](http://www.github.com/Robinovelace/Creating-maps-in-R). Run the commands from the [unzipped folder](https://github.com/Robinlovelace/Creating-maps-in-R/archive/master.zip) to see the commands in action!

Using the example of the `help` command, type:
`?help` (for information on a specific function),
`??help` (to search for word in R's documentation)
and `example(help)` (for pre-made examples of that function).

**Key packages**

Additional packages must be installed (using `install.packages('packageName')`) on top of 
R's [default packages](http://www.hep.by/gnu/r-patched/r-admin/R-admin_33.html) to utilise R's spatial 
capabilities. Packages are loaded using `library(packageName)`. For example, `r library(rgdal)` will load the **rgdal** package which is R's interface to the 'Geospatial Abstraction Library' (GDAL). The most important and frequently used *general purpose* spatial packages are:

- **sp**: basis of most other spatial R packages - provides Spatial* classes
- **rgdal**: R's interface to the GDAL data import library
- **rgeos**: provides a number of spatial functions
- **maptools**: tools for handling spatial objects

**raster** and **spatstat** are mature and extremely useful spatial packages, providing functions for raster and point-pattern applications, respectively.
There are hundreds of additional R packages described on the [CRAN website](http://cran.r-project.org) in the [*spatial view*](http://cran.r-project.org/web/views/Spatial.html).

# R's Spatial* classes

To work with spatial data, R uses *classes* that are more complex
than the default data classes in R's base package; *S3* `vector`, `list` and `data.frame` s. 
Spatial object classes can be identified with the
function `class(objectName)`. Each contains a number of different data
*slots*: sub-classes of specific types. Some important
spatial classes provided by the `sp` package include:

- `SpatialPoints`: point data, each point represented by an x and y coordinate
- `SpatialLines`: line data, composed of lists of the `Lines` sub-class or *slot*
- `SpatialPolygons`: polygon data, with `Polygons` and `Polygon` *slots* dewwed
- `SpatialPixels`: a spatial class for raster data comprised of pixels

It is important to note that each of the above classes cannot contain non-spatial
attribute data (like attribute tables in GIS softward like QGIS). Therefore, they must first be converted into a class of the same name but suffixed with `DataFrame`. Thus `london <- SpatialPointsDataFrame(sp, df)` will create a new object containing points with spatial and non-spatial information called london. Likewise `SlDf <- SpatialLinesDataFrame` will create a new object containing lines with spatial and non-spatial information. 

The attribute data of the london object can be accessed in this instance by using `spPdf@data`. The `@data` notation refers to the `data` slot of the new object. More fundamental spatial classes, which
are in fact subsets of the `Spatial*` type classes listed above, include `bbox` (the
object's bounding box) and `proj4string` (the projection of the object, which may be `NA`).

**Loading spatial data**
`readOGR` from **rgdal** can load a wide variety of spatial data types. The following line loads spatial data representing London:

```{r, warning=FALSE, message=FALSE, results='hide'}
london <- readOGR("data/", "london_sport") 
```

Plot the london object using the geometry slot:

plot(london)

#Analysing Attribute data

Print the column headings of the london object:
```{r}
names(london)
```

Print the entire london object:
```{r}
print(london)
```

Print a defined number of rows from the london object (2 in this case):
```{r}
head(london@data, n=2)
```

Some basic column statistics:
```{r}
mean(london$Partic_Per) #calculate the mean value of the Partic_Per column
max(london$Partic_Per)  #calculate the maximum value 
min(london$Partic_Per)  #calculate the minimum value
sum(london$Partic_Per)  #Calculate the sum of the values

nrow(london)  #Return the number of rows in the dataset
ncol(london)  #Return the number of columns in the dataset 
```

Additional information about the london object:
```{r}
summary(london)
```

# Allocating and changing projection
Before undertaking any further analysis it is useful to know the Coordinate Reference System(CRS) of the london object. 

Query the current projection of an object: 
```{r}
proj4string(london)
```
Change the Coordinate Reference System (CRS) if it has been incorrectly assigned:

Change the Coordinate Reference System (CRS) if it has been incorrectly assigned:
```{r}
proj4string(london) <-CRS("+init=epsg:27700")
```

Note the warning above which states that the coordinate reference system has been changed but the data
has not been transformed. To transform the data use:

```{r}
london.wgs84 <-spTransform(london, CRS("+init=epsg:4326"))
```

#Attribute Joins

Attribute joins are used to link additional pieces of information to pre-existing polygons. In the london object, for example, there are 5 attribute variables (you saw these earlier when typing `names(london)`)

# Spatial subsetting



# Spatial aggregation

# Spatial graphics

# Basemaps and advanced functions



