---
title: "R Spatial Reference Card"
author: "Robin"
date: "10/07/2014"
output: pdf_document
---

R has a suite of powerful and effective spatial data processing tools. It is the purpose
of this reference card to summarise the most important and frequently used
packages and functions. Some of the commands use real datasets from the
'Creating-maps-in-R' [GitHub repository](http://www.github.com/Robinovelace/Creating-maps-in-R). Unzip this repo and run the commands from the folder's root directory to see the commands in action, or use R's help functions.

Using the example of the `help` command, type:
`?help` (for information on a specific function),
`??help` (to search for word in R's documentation)
and `example(help)` (for pre-made examples of that function).

# Key packages

Additional packages must be installed (using `install.packages('packageName')`) on top of 
R's [default packages](http://www.hep.by/gnu/r-patched/r-admin/R-admin_33.html) to utilise R's spatial 
capabilities. Packages are loaded using `library(packageName)`.For example: the `library(rgdal)` will load the rgdal package which is R's interface to the 'Geospatial Abstraction Library(GDAL). rgdal has a number of useful functions including `readOGR` which loads spatial data formats such as shapefiles.  

```{r}
library(rgdal)
```

Some other important and frequently used spatial packages are:

- **sp**: basis of most other spatial R packages - provides Spatial* classes
- **rgdal**: R's interface to the GDAL data import library
- **rgeos**: provides a number of spatial functions
- **maptools**: tools for handling spatial objects
- **raster**: provides functions for raster data

There are hundreds of additional R packages described on the Comprehensive
R Archive Network's (CRAN) website under the
[*spatial view*](http://cran.r-project.org/web/views/Spatial.html).

# R's Spatial* classes

To work with spatial data, R uses *classes* that are more complex
than the default data classes in R's base package; *S3* `vector`, `list` and `data.frame` s. 
Spatial object classes can be identified with the
function `class(objectName)`. Each contains a number of different data
*slots*: sub-classes of specific types. Some important
spatial classes provided by the `sp` package include:

- `SpatialPoints`: point data, each point represented by an x and y coordinate
- `SpatialLines`: line data, composed of lists of the `Lines` sub-class or *slot*
- `SpatialPolygons`: polygon data, with `Polygons` and `Polygon` *slots* dewwed
- `SpatialPixels`: a spatial class for raster data comprised of pixels

It is important to note that each of the above classes cannot contain non-spatial
attribute data (imagine an attribute table in a GIS). Therefore, they must first be converted into a class of the same name but suffixed with `DataFrame`. Thus `SpDf <- SpatialPointsDataFrame(sp, df)` will create a new object containing points with spatial and non-spatial information called SpDf. Likewise `SlDf <- SpatialLinesDataFrame` will create a new object containing lines with spatial and non-spatial information. 

The attribute data of the SpDf object can be accessed in this instance by using `spPdf@data`. The `@data` notation refers to the `data` slot of the new object. More fundamental spatial classes, which
are in fact subsets of the `Spatial*` type classes listed above, include `bbox` (the
object's bounding box) and `proj4string` (the projection of the object, which may be `NA`).

# Loading spatial data
Create a new spatial object called SpDf using the 'london_sport' shapefile: 
```{r, warning=FALSE, message=FALSE, results='hide'}
SpDf <- readOGR("data/", "london_sport") 
```

Plot the SpDf object using the geometry slot:

plot(SpDf)

#Analysing Attribute data

Print the column headings of the SpDf object:
```{r}
names(SpDf)
```

Print the entire SpDf object:
```{r}
print(SpDf)
```

Print a defined number of rows from the SpDf object (2 in this case):
```{r}
head(SpDf@data, n=2)
```

Some basic column statistics:
```{r}
mean(SpDf$Partic_Per) #calculate the mean value of the Partic_Per column
max(SpDf$Partic_Per)  #calculate the maximum value 
min(SpDf$Partic_Per)  #calculate the minimum value
sum(SpDf$Partic_Per)  #Calculate the sum of the values

nrow(SpDf)  #Return the number of rows in the dataset
ncol(SpDf)  #Return the number of columns in the dataset 
```

Additional information about the SpDf object:
```{r}
Summary(SpDf)
```

# Allocating and changing projection
Before undertaking any further analysis it is useful to know the Coordinate Reference System(CRS) of the SpDf object. 

Query the current projection of an object: 
```{r}
proj4string(SpDf)
```

Change the Coordinate Reference System (CRS) if it has been incorrectly assigned:
```{r}
proj4string(SpDf) <-CRS("+init=epsg:27700")
```

Note the warning above which states that the coordinate reference system has been changed but the data
has not been transformed. To transform the data use:
```{r}
SpDf.wgs84 <-spTransform(SpDf("+init=epsg:4326"))
```

#Attribute Joins
Attribute joins are used to link additional pieces of information to pre-existing polygons. In the SpDf object, for example, there are 5 attribute variables (you saw these earlier when typing `names(SpDf)`)

# Spatial subsetting



# Spatial aggregation

# Spatial graphics

# Basemaps and advanced functions



