---
title: "Basic mapping and attribute joins in R"
author: "Robin Lovelace"
output:
  html_document:
    keep_md: yes
    self_contained: no
---

This post is based on the free and open source [Creating-maps-in-R](https://github.com/Robinlovelace/Creating-maps-in-R) created for teaching R as a command-line GIS.

R is well known as an language ideally suited for
data processing, statistics and modelling. R has a number of spatial packages,
allowing analyses that would require hundreds of lines of code in other languages
to be implemented with relative ease. Geographically weighted regression,
analysis of space-time data and raster processing are three niche
areas where R outperform much of the competition, thanks to community contributions such as
**[spgwr](http://cran.r-project.org/web/packages/spgwr/index.html)**, 
**[spacetime](http://cran.r-project.org/web/packages/spacetime/index.html)**
and the wonderfully straightforward **[raster](http://cran.r-project.org/web/packages/raster/index.html)** packages.

What seems to be less well known is that R performs
well as a self standing Geographical Information System (GIS) in its own right.
Everyday tasks such as reading and writing geographical data formats, reprojecting,
joining, subsetting and overlaying spatial objects can be easy and intuitive in
R, once you understand the slightly specialist data formats and syntax of spatial
R objects and functions. These basic operations are the basic foundations of GIS. Mastering
them will make much more advanced operations much easier. Based on the saying
'master walking before trying to run', this mini tutorial demonstrates how to
load and plot a simple geographical object in R, illustrating that the ease with
which continuous and binned choropleth map color schemes can be created using
**[ggmap](http://cran.r-project.org/web/packages/ggmap/index.html)**, an extension
of the popular
**[ggplot2](http://cran.r-project.org/package=ggplot2)** graphics package.
Crucially, we will also see how to *join* spatial and non spatial datasets,
resulting in a map of where the Conservative party succeeded and failed
in gaining council seats in the
[2014 local elections](http://en.wikipedia.org/wiki/United_Kingdom_local_elections,_2014).

<--more-->

As with any project, the starting point is to load the data we'll be using.
In this case we can download all the datasets from a single souce:
the [Creating-maps-in-R](https://github.com/Robinlovelace/Creating-maps-in-R)
github repository which is designed to introduce R's basic geographical
functionality to beginners. We can use R to download and unzip
the files using the following commands (from a Linux-based
[operating system](http://lubuntu.net/)). This ensures reproducibility:

```{r, message=FALSE, results='hide'}
# load the packages we'll be using for this tutorial
x <- c("rgdal", "dplyr", "ggmap", "RColorBrewer")
lapply(x, library, character.only = TRUE)
```


```{r, eval=FALSE}
# download the repository:
download.file("https://github.com/Robinlovelace/Creating-maps-in-R/archive/master.zip", destfile = "rmaps.zip", method = "wget")
unzip("rmaps.zip") # unzip the files
```

Once 'in' the folder, R has easy access to all the datasets we 
need for this tutorial. As this is about GIS, the first stage is to
load and plot some spatial data: a map of London:

```{r, message=FALSE}
setwd("/home/robin/Desktop/Creating-maps-in-R-master/") # navigate into the unzipped folder
london <- readOGR("data/", layer = "london_sport")
plot(london)
```

The data has clearly loaded correctly and can be visualised, but where
*is* it? The `london` is simply printed, a load of unreadable information
is printed, including the coordinates defining the geographical extent
of each zone and additional non-geographical attributes.
The [polymophic](http://heather.cs.ucdavis.edu/~matloff/R/RProg.pdf)
means that [generic functions](http://adv-r.had.co.nz/OO-essentials.html#s3)
behave differently depending on the type of data they are fed.
The following command, for example, is actually calling `mean.Date`
behind the scenes, allowing R to tell us that the the 2^nd^ of July
was half way through the year. The default `mean.default`
function does not work:

```{r}
mean(as.Date(c("01/01/2014", "31/12/2014"), format = "%d/%m/%Y"))
```

In the same way, we can use the trusty `summary` function to summarise
our R object:

```{r}
summary(london)
```

This has outputed some very useful information: the bounding box
of the object, its coordinate reference system (CRS) and even
summaries of the attributes associated with each zone. `nrow(london)`
will tell us that there are 33 polygons represented within the object.

To gain a fuller understanding of the structure of the `london` object,
we can use the `str` function (but only on the first polygon, to avoid
an extrememly long output):

```{r}
str(london[1,])
```

This shows us that the fundamental structure of a `SpatialPolygonsDataFrame`
is actually rather complicated. This complexity is useful, allowing R to store
the full range of information needed to describe almost any polygon-based dataset.
The `@` symbol in the structure represents *slots* which
are specific to the
[S4 object class](http://adv-r.had.co.nz/OO-essentials.html#s4) and
contain specific
pieces of information within the wider `london` object. The basic
slots within the london object are:

- `@data`, which contains the the attribute data for the zones
- `@polygons`, the geographic data associated with each polygon (this confusingly contains the `@Polygons` slot: each polygon feature can contain multiple `Polygons`, e.g. if an administrative zone is non-contiguous)
- `@plotOrder` is simply the order in which the polygons are plotted
- `@bbox` is a slot associated with all spatial objects, representing its spatial extent
- `@proj4string` the CRS associated with the object

Critically for exploring the attributes of `london` is the `data` slot.
We can look at and modify the attributes of the subdivisions of `london` easily
using the `@` notation:

```{r}
head(london@data)
```

Having seen his notation, many (if not most) R beginners will tend to
always use it to refer to attribute data in spatial objects. Yet `@` is often not
needed. To refer to the population of London, for example, the following lines
of code yield the same result:

```{r}
mean(london@data$Pop_2001)
mean(london$Pop_2001)
```

Thus we can treat the S4 spatial data classes as if they were regular
data frames in *some* contexts, which is extremely useful for concise code.
To plot the population of London zones on a map, the following code
[works](http://www.stat.ubc.ca/~jenny/STAT545A/block14_colors.html):

```{r}
cols <- brewer.pal(n = 4, name = "Greys")
lcols <- cut(london$Pop_2001,
  breaks = quantile(london$Pop_2001),
  labels = cols)
plot(london, col = as.character(lcols))
```

Now, how about joining additional variables to the spatial object?
To join information to the existing variables, the *join* functions
from **dplyr** (which replaces and improves on **plyr**) are a godsend.
The following code loads a non-geographical dataset and joins an additional
variable to `london@data`:

```{r}
ldat <- read.csv("/home/robin/Desktop/Creating-maps-in-R-master/data/london-borough-profiles-2014.csv")
dat <- select(ldat, Code, contains("Anxiety"))
dat <- rename(dat, ons_label = Code, Anxiety = Anxiety.score.2012.13..out.of.10.)
dat$Anxiety <- as.numeric(as.character(dat$Anxiety))
london@data <- left_join(london@data, dat)
head(london@data) # the new data has been added
```

## Plotting maps with **ggplot**

In order to plot the average anxiety scores across london we can use **ggplot2**:

```{r}
lf <- fortify(london, region = "ons_label")
lf <- rename(lf, ons_label = id)
lf <- left_join(lf, london@data)
ggplot(lf) + geom_polygon(aes(long, lat, group = group, fill = Anxiety))
```

## The challenge

Using the skills you have learned in the above tutorial, see if you can
replicate the graph below: the proportion of Conservative councilors selected
in different parts of London. Hint: the data is contained in `ldat`, as
downloaded from here: http://data.london.gov.uk/dataset/london-borough-profiles.

```{r, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
dat <- rename(ldat, conservative = Proportion.of.seats.won.by.Conservatives.in.2014.election)
dat$conservative <- as.numeric(as.character(dat$conservative))
summary(dat$conservative)
head(london$ons_label)
head(dat$Code)
# check the joining variables work
summary(dat$Code %in% london$ons_label)
# rename the linking variable
dat <- rename(dat, ons_label = Code)
to_join <- select(dat, ons_label, conservative)
london@data <- left_join(london@data, to_join)

london_f <- fortify(london, region = "ons_label") # you may need to load maptools
london_f <- rename(london_f, ons_label = id)

london_f <- left_join(london_f, london@data)

map <- ggplot(london_f, aes(long, lat, group = group, fill = conservative)) +
  geom_polygon() +
  coord_equal() +
  labs(x = "Easting (m)", y = "Northing (m)", fill = "% who\nvoted\nConservative") +
  scale_fill_continuous(low = "grey", high = "blue") +
  theme_nothing(legend = T)
map
```










