---
title: "Clipping spatial data with spatial objects in R"
author: "Robin Lovelace"
date: "29/07/2014"
output: 
  md_document:
    variant: markdown_github
---

This miniature vignette shows how to clip spatial data based
on different spatial objects in R and a 'bounding box'. Spatial
overlays are common in GIS applications and R users are fortunate
that the clipping and spatial subsetting functions are mature and
fairly fast. We'll also write a new function called `gClip()`,
that will make clipping by bounding boxes easier.

## Loading the data

To start with let's load some data. I'm working in the root directory
of the [Creating-maps-in-R](https://github.com/Robinlovelace/Creating-maps-in-R)
github repository, so there are some spatial datasets available to play with:

```{r, warning=FALSE, message=FALSE}
setwd("../")
library(rgdal)
stations <- readOGR("data", "lnd-stns")
zones <- readOGR("data", "london_sport")
```

## The wonder of spatial subsetting in R

Now, it's easy to **subset** spatial data in R, using the
same incredibly concise square bracket `[]` notation as R uses for non spatial data.
To re-confirm how this works on non-spatial data, here's a mini example:

```{r}
M <- matrix(1:10, ncol = 5)
M[2, 3:5]
```

The above code creates a matrix with 5 columns and 2 rows: the `[2, 3:5]` part
takes the subset of `M` corresponding to 3rd to 5th columns in the second row.

Subsetting spatial data works in exactly the same way: note that `zones` are
far more extensive than the `stations` points. (We have to change the projection
of `stations` before plotting, so the objects are on the same coordinate reference system.)

```{r, warning=FALSE}
stations <- spTransform(stations, CRS(proj4string(zones))) # transform CRS
plot(zones)
points(stations)
```

So how do we select only points that are are within the zones of London?
Believe it or not, it's as simple as subsetting the matrix `M` above.
This is an amazingly concise and convenient way of spatial subsetting
that was added to R at some point between versions 1 and 2 of 
[Applied Spatial Data Analysis with R](http://www.asdar-book.org/).
In the earlier (2008) book, one had to use `overlay(x, y)` just to
get the selection, and then another line of code was required
to actually make the subset. Now things are much simpler.
As Bivand et al. put it in the
[latest edtion](http://www.springer.com/statistics/life+sciences,+medicine+%26+health/book/978-1-4614-7617-7) (p. 131),
"the selection syntax for features was estended such that it understands:"

```{r}
stations_subset <- stations[zones, ]
```

**This is so amazing and intuitive, whover invented it should be given a medal!!**
Despite this simplicity, it seems many R users
who I've taugh spatial functions to are unaware of `[]`'s spatial extension.
So spread the word (and if anyone knows of the history of this innovation, please let us know).
Now we have a sample of all stations zones: progress.

```{r}
plot(zones)
points(stations_subset)
```

## Clipping by a bounding box

But what if we want to *clip* the polygons data, based on a bounding box?
To start with, let's look at and modify the existing bounding box for the zones, making it
half the size:

```{r}
b <- bbox(zones)
b[1, ] <- (b[1, ] - mean(b[1, ])) * 0.5 + mean(b[1, ])
b[2, ] <- (b[2, ] - mean(b[2, ])) * 0.5 + mean(b[2, ])
b <- bbox(t(b))
plot(zones, xlim = b[1, ], ylim = b[2, ])
```

Now, to clip this area, we can use a custom function, which I've called
`gClip`, following the **rgeos** function naming convention (this was inspired
by an online [answer](http://stackoverflow.com/questions/21883683/is-it-possible-to-clip-a-polygon-to-the-bounding-box-of-a-base-map) that didn't work for me):

```{r}
library(raster)
library(rgeos)
gClip <- function(shp, bb){
  if(class(bb) == "matrix") b_poly <- as(extent(as.vector(t(bb))), "SpatialPolygons")
  else b_poly <- as(extent(bb), "SpatialPolygons")
  gIntersection(shp, b_poly, byid = T)
}

zones_clipped <- gClip(zones, b)
plot(zones_clipped)
```

Note that due to the `if` statements in `gClip`'s body, it can handle almost
any spatial data input, and still work. Let's clip to the borough of
Westminster, one of London's better known boroughs:

```{r Westminster}
westminster <- zones[grep("West", zones$name),]
zones_clipped_w <- gClip(zones, westminster)
plot(zones_clipped_w); plot(westminster, col = "red", add = T)
```

## Conclusion

I hope this is post has been useful. If so, there are many more spatial tips available
from the [Introduction to visualising spatial data in R](https://github.com/Robinlovelace/Creating-maps-in-R/raw/master/intro-spatial-rl.pdf) that [James Cheshire](http://spatial.ly/).
The source code of this post can also be viewed [online](https://github.com/Robinlovelace/Creating-maps-in-R/blob/master/vignettes/clipping-with-bounding-box.Rmd) as just one of a series of
[vignettes](https://github.com/Robinlovelace/Creating-maps-in-R/tree/master/vignettes)
to showcase some of R's impressive spatial capabilities.
