---
title: "Reading and writing GeoJSON files in R"
author: "Robin Lovelace"
date: "11/26/2014"
output: html_document
---

[GeoJSON](http://geojson.org/) is an open data format that
extends [JSON](http://www.json.org/), the JavaScript Object Notation
to encode spatial information.

GeoJSON has the following advantages:

- All the geographical information is self contained (unlike shapefiles, which have multiple files for one spatial object)
- It is an open standard which is becoming increasingly popular
- GeoJSON files display as maps on GitHub by default!

The disadvantages are:

- GeoJSON files are [quite large](http://zevross.com/blog/2014/04/22/spatial-data-on-a-diet-tips-for-file-size-reduction-using-topojson/) compared with non text formats such as shapefiles and
spatial databases.
- GeoJSON files can be slow to read/write and are not supported by
all GIS software (yet!)

As illustrated by this [post](http://www.jose-gonzalez.org/create-geojson-r/#.VHVoMY_6IrE), it is easy to read and write GeoJSON files
in R, using the [**rgdal**](http://cran.r-project.org/web/packages/rgdal/index.html) package, which calls
[gdal](http://www.gdal.org/) commands. Here's an example using
publicly available GeoJSON file:

```{r, echo=FALSE}
system("rm /tmp/gas*")
```


```{r}
library(rgdal) # the library we'll be using
u <- "http://giscience.tw.rpi.edu/rpi_group_projects/watershedders/data/gas_station.geojson"
download.file(url = u, destfile = "/tmp/gas.GeoJSON")
gas <- readOGR(dsn = "/tmp/gas.GeoJSON", layer = "OGRGeoJSON")
summary(gas)
```

Note that to test whether the GeoJSON driver works on your system,
you can use the following [command](http://stackoverflow.com/questions/24183007/is-it-possible-to-read-geojson-or-topojson-file-in-r-to-draw-a-choropleth-map):

```{r}
"GeoJSON" %in% ogrDrivers()$name
```

As shown in the above example,
GeoJSON files can have coordinate reference system (CRS)
allocated to them through the `"crs"` 'member':

```{r, engine='node', eval=FALSE}
  "crs": {
    "type": "name",
    "properties": {
      "name": "urn:ogc:def:crs:OGC:1.3:CRS84"
      }
    }
```

Note that we can see this definition in the `gas.GeoJSON` file:

```{r}
readLines("/tmp/gas.GeoJSON")[3]
```


This 'member' is defined in plain text in the GeoJSON file
before the spatial `"features"` are defined. Please see this
small example of [gas stations](http://giscience.tw.rpi.edu/rpi_group_projects/watershedders/data/gas_station.geojson) to see how the CRS
is defined in the context of a complete GeoJSON file.

To write GeoJSON files, the sensibly named corrollary of
`readOGR` is used, `writeOGR`:

```{r, echo=FALSE}
file.remove("/tmp/gas2.GeoJSON")
```


```{r}
writeOGR(gas, dsn = "/tmp/gas2.GeoJSON", layer = "u", driver = "GeoJSON")
```

Let's see what R has created:

```{r}
gas2 <- readLines("/tmp/gas2.GeoJSON")
gas2[1:4]
```

Frustratingly, this fails to write the CRS into the file.
**rgdal**'s ability to read a crs from a GeoJSON file is no
guarantee of it's ability to then write it out again.
**rgdal** has **lost the CRS in translation**.

This can problem be caused by using an old version of
`gdal`. Not in this case though: incomplete CRS values can
lead `rgdal` to omit the `"crs"` in the file, as explained by
Roger Bivand in a lengthy [conversation](http://r-sig-geo.2731867.n2.nabble.com/WriteOGR-to-GeoJSON-loses-CRS-td7586913.html)
on the [R-sig-geo](https://stat.ethz.ch/mailman/listinfo/r-sig-geo)
email list. To test that the problem is the fault of gdal and
not R, we can use the `ogr2ogr` command line tool, supplied by
gdal. 

```{r, echo=FALSE}
system("rm /tmp/gas3.*")
```


```{r}
# use call gdal from the operating system to convert original GeoJSON:
system("ogr2ogr -f 'ESRI Shapefile' /tmp/gas3.shp '/tmp/gas.GeoJSON' ")

gas3 <- readOGR(dsn = "/tmp", layer = "gas3")
proj4string(gas3)
proj4string(gas)
```

The above code shows that gdal has successfully converted the
original GeoJSON file to a shapefile, maintaining the CRS.
(Note that the CRS is identical but its definition is slightly different.)
Let's take a look at the `.proj` file that gdal created:

```{r}
gas3_proj <- readLines("/tmp/gas3.prj", skipNul = T)
gas3_proj
```

# gdal fails to convert certain types of CRS

There are various ways to solve the problem.
The 'outside R' solution would be to write the file
first as a shapefile, and then use `ogr2ogr` to convert
it to a GeoJSON. But as shown in the code below, this fails:

```{r}
writeOGR(gas, "/tmp/", layer = "gas4", driver = "ESRI Shapefile")
system("ogr2ogr -f 'GeoJSON' /tmp/gas5.GeoJSON '/tmp/gas4.shp' ")
readLines("/tmp/gas4.prj")
readLines("/tmp/gas5.GeoJSON")[1:3]
```

# CRS definitions

As Roger Bivand shows, if the CRS is defined in certain
ways, the `"crs"` member will be written to the GeoJSON file.
But it is difficult to know how to define a CRS other than
by its [EPSG name](http://spatialreference.org/ref/epsg/4283/):

```{r}
gas6 <- gas # duplicate the gas object
proj4string(gas6) <- CRS("+init=epsg:4283")
proj4string(gas6)
# the below works, but incorrect CRS
# proj4string(gas6) <- CRS("+proj=longlat +ellps=clrk66 +datum=NAD27")
writeOGR(gas6, dsn = "/tmp/gas6.GeoJSON", layer = "gas", driver = "GeoJSON")
readLines("/tmp/gas6.GeoJSON")[1:3]
```

# A manual solution

To solve the problem manually, we can simply add the correct projection
to the GeoJSON file that has been created. Thankfully this is straightforward, as line 3 of the files are left empty by gdal, presumably ready
for such an eventuality:

```{r}
gas7 <- readLines("/tmp/gas2.GeoJSON")
gas7[1:4]
gas7[3] <- '"crs": {"type": "EPSG", "properties": {"code": 4283}},'
writeLines(gas7, con = "/tmp/gas8.GeoJSON")
gas8 <- readOGR(dsn = "/tmp/gas8.GeoJSON", layer = "OGRGeoJSON")
proj4string(gas8)
```

This is a messy solution but it's the only one I could find to write
GeoJSON files with a CRS defined for Australian data services' default CRS.










```{r, echo=FALSE}
# library(rgdal)
# library(pryr)
# pryr::show_c_source(readOGR)
# showMethods(readOGR)
# ??readOGR
```


