---
title: "Reading and writing GeoJSON files in R"
author: "Robin Lovelace"
date: "11/26/2014"
output: html_document
---

## Introduction

[GeoJSON](http://geojson.org/) is an open data format that
extends [JSON](http://www.json.org/), the JavaScript Object Notation
to encode spatial information.

GeoJSON has the following advantages:

- All the geographical information is self contained (unlike shapefiles, which have multiple files for one spatial object)
- It is an open standard which is becoming increasingly popular
- GeoJSON files display as maps on GitHub by default!

The disadvantages are:

- GeoJSON files are [quite large](http://zevross.com/blog/2014/04/22/spatial-data-on-a-diet-tips-for-file-size-reduction-using-topojson/) compared with non text formats such as shapefiles and
spatial databases.
- GeoJSON files can be slow to read/write and are not supported by
all GIS software (yet!).

## Reading GeoJSONs with readOGR

As illustrated by this [post](http://www.jose-gonzalez.org/create-geojson-r/#.VHVoMY_6IrE), it is easy to read and write GeoJSON files
in R, using the [**rgdal**](http://cran.r-project.org/web/packages/rgdal/index.html) package, which calls
[gdal](http://www.gdal.org/) commands. Here's an example using
publicly available GeoJSON file:

```{r, echo=FALSE}
system("rm /tmp/gas*")
```


```{r}
library(rgdal) # the library we'll be using
u <- "https://raw.githubusercontent.com/codeforboston/open_data_cambridge/master/Infrastructure/UtilityFacilities.geojson"
downloader::download(url = u, destfile = "/tmp/gas.GeoJSON")
gas <- readOGR(dsn = "/tmp/gas.GeoJSON", layer = "OGRGeoJSON")
summary(gas)
```

Note that to test whether the GeoJSON driver works on your system,
you can use the following [command](http://stackoverflow.com/questions/24183007/is-it-possible-to-read-geojson-or-topojson-file-in-r-to-draw-a-choropleth-map):

```{r}
"GeoJSON" %in% ogrDrivers()$name
```

As shown in the above example,
GeoJSON files can have coordinate reference system (CRS)
allocated to them through the `"crs"` 'member':

```{r, engine='node', eval=FALSE}
  "crs": {
    "type": "name",
    "properties": {
      "name": "urn:ogc:def:crs:OGC:1.3:CRS84"
      }
    }
```

Note that we can see this definition in the `gas.GeoJSON` file:

```{r}
readLines("/tmp/gas.GeoJSON")[3]
```


This 'member' is defined in plain text in the GeoJSON file
before the spatial `"features"` are defined. Please see this
small example of [gas stations](http://giscience.tw.rpi.edu/rpi_group_projects/watershedders/data/gas_station.geojson) to see how the CRS
is defined in the context of a complete GeoJSON file.

## Writing GeoJSONs with writeOGR

To write GeoJSON files, the sensibly named corrollary of
`readOGR` is used, `writeOGR`:

```{r}
if(file.exists("/tmp/gas2.geojson")){
  file.remove("/tmp/gas2.geojson")
}
```


```{r}
geojsonio::geojson_write(gas, file = "/tmp/gas2.geojson")
```

## Multifeature geojsons

Based on [this](https://stackoverflow.com/questions/30583048/convert-features-of-a-multifeature-geojson-into-r-spatial-objects/30593077#30593077), geojson files can usually be loaded into R with `readOGR`, as illustrated above.

However, this fails for multifeature geojsons.

Reproducible example:

```{r}
downloader::download("https://github.com/Robinlovelace/Creating-maps-in-R/raw/master/data/test-multifeature.geojson", "test.geojson")
try(test <- rgdal::readOGR("test.geojson", "OGRGeoJSON")) # fails with:
```

There are 2 solutions. One uses rgdal and the gdal argument
`require_geomType`. The other uses geojsonio.

### Multifeature geojsons with rgdal

The first solution uses new features introduced in gdal 0.9-3.
It is worth checking (and maybe updating) your gdal version.
(Thanks to [@hrbrmstr](https://stackoverflow.com/users/1457051/hrbrmstr) for this detailed information.)

```{r}
rgdal::getGDALVersionInfo() # check the version of gdal
system("gdal-config --version") # asking gdal its version from bash
```

```
Before update, this was:
[1] "GDAL 1.11.1, released 2014/09/24"

After updating gdal, e.g. like this (Ubuntu), it changed:
$ sudo add-apt-repository ppa:ubuntugis/ubuntugis-unstable && sudo apt-get update
$ sudo apt-get install gdal-bin
```

gdal can provide us with information about the layers in the
multipart geojson.

```{r}
ogrListLayers("test.geojson")
```

It can also provide more general information about the object,
if we ask it correctly.

```{r}
try(ogrInfo("test.geojson")) # no information on the file
system("ogrinfo test.geojson ")
try(ogrInfo("test.geojson", "OGRGeoJSON"))
```

The output from the previous lines indicates how we can
access just one layer. We use the argument `require_geomType`
to specify which of the geometries to load:

```{r}
testp_info <-
  ogrInfo("test.geojson", "OGRGeoJSON", require_geomType="wkbPoint")
testp_df <- as.data.frame(testp_info$iteminfo)
head(testp_df)
```

Note we can use the same command for any of the geometries:

```{r, eval=FALSE}
ogrInfo("test.geojson", "OGRGeoJSON", require_geomType = "wkbPolygon")
```
```
NOTE: keeping only 23 wkbPolygon of 781 features
    note that extent applies to all features
Source: "test.geojson", layer: "OGRGeoJSON"
Driver: GeoJSON number of rows 781 
  selected geometry type: wkbPolygon with 23 rows
Feature type: wkbPolygon with 2 dimensions
Extent: (12.48326 41.88355) - (12.5033 41.89629)
CRS: +proj=longlat +datum=WGS84 +no_defs  
Number of fields: 78 
                       name type length typeName
1                      area    4      0   String
2                   bicycle    4      0   String
3                   highway    4      0   String
```

Armed with this information, we can read in the features:

```{r}
points <- readOGR("test.geojson", "OGRGeoJSON", require_geomType="wkbPoint")

lines <- readOGR("test.geojson", "OGRGeoJSON", require_geomType="wkbLineString")

polygons <- readOGR("test.geojson", "OGRGeoJSON", require_geomType="wkbPolygon")
```

Let's plot the data (see https://stackoverflow.com/questions/30583048/ )

```{r}
plot(lines, col="#7f7f7f")
plot(polygons, add=TRUE)
plot(points, add=TRUE, col="red")
```

## Solution from geojsonio

This is an alternative solution:

```{r}
test <- geojsonio::geojson_read("test.geojson")
```

Then we can filter out the feature types of interest:

```{r}
obj = Filter(
    function(f){f$geometry$type=="LineString"},
    test$features)
```

## Solution from Spacedman

```{r}
mess <- 'ogr2ogr -where "OGR_GEOMETRY=\'LINESTRING\'" -f "GeoJSON" lines.geojson  test.geojson'
system(mess)
testl <- rgdal::readOGR("lines.geojson", layer = "OGRGeoJSON")
```


  [1]: https://github.com/Robinlovelace/Creating-maps-in-R/blob/master/vignettes/geoJSON.Rmd
  [2]: https://github.com/Robinlovelace/Creating-maps-in-R/blob/master/data/test-multifeature.geojson


## gdal fails to convert certain types of CRS

Let's see what R has created from writeOGR:

```{r}
gas2 <- readLines("/tmp/gas.GeoJSON")
gas2[1:4]
```

Frustratingly, this fails to write the CRS into the file.
**rgdal**'s ability to read a crs from a GeoJSON file is no
guarantee of it's ability to then write it out again.
It seems
**rgdal** has **lost the CRS in translation**.

This can problem be caused by using an old version of
`gdal`. Not in this case though: incomplete CRS values can
lead `rgdal` to omit the `"crs"` in the file, as explained by
Roger Bivand in a lengthy [conversation](http://r-sig-geo.2731867.n2.nabble.com/WriteOGR-to-GeoJSON-loses-CRS-td7586913.html)
on the [R-sig-geo](https://stat.ethz.ch/mailman/listinfo/r-sig-geo)
email list. To test that the problem is the fault of gdal and
not R, we can use the `ogr2ogr` command line tool, supplied by
gdal. 

```{r, echo=FALSE}
system("rm /tmp/gas3.*")
```


```{r}
# use call gdal from the operating system to convert original GeoJSON:
system("ogr2ogr -f 'ESRI Shapefile' /tmp/gas3.shp '/tmp/gas.GeoJSON' ")

gas3 <- readOGR(dsn = "/tmp", layer = "gas3")
proj4string(gas3)
proj4string(gas)
```

The above code shows that gdal has successfully converted the
original GeoJSON file to a shapefile, maintaining the CRS.
(Note that the CRS is identical but its definition is slightly different.)
Let's take a look at the `.proj` file that gdal created:

```{r}
gas3_proj <- readLines("/tmp/gas3.prj", skipNul = T)
gas3_proj
```

There are various ways to solve the problem.
The 'outside R' solution would be to write the file
first as a shapefile, and then use `ogr2ogr` to convert
it to a GeoJSON. But as shown in the code below, this fails:

```{r}
writeOGR(gas, "/tmp/", layer = "gas4", driver = "ESRI Shapefile")
system("ogr2ogr -f 'GeoJSON' /tmp/gas5.GeoJSON '/tmp/gas4.shp' ")
readLines("/tmp/gas4.prj")
readLines("/tmp/gas5.GeoJSON")[1:3]
```

# CRS definitions

As Roger Bivand shows, if the CRS is defined in certain
ways, the `"crs"` member will be written to the GeoJSON file.
But it is difficult to know how to define a CRS other than
by its [EPSG name](http://spatialreference.org/ref/epsg/4283/):

```{r}
gas6 <- gas # duplicate the gas object
proj4string(gas6) <- CRS("+init=epsg:4283")
proj4string(gas6)
# the below works, but incorrect CRS
# proj4string(gas6) <- CRS("+proj=longlat +ellps=clrk66 +datum=NAD27")
geojsonio::geojson_write(gas6, file = "/tmp/gas6.geojson")
readLines("/tmp/gas6.geojson")[1:4]
```

# A manual solution

To solve the problem manually, we can simply add the correct projection
to the GeoJSON file that has been created. Thankfully this is straightforward, as line 3 of the files are left empty by gdal, presumably ready
for such an eventuality:

```{r}
gas7 <- readLines("/tmp/gas2.geojson")
gas7[1:4]
gas7[3] <- '"crs": {"type": "EPSG", "properties": {"code": 4283}},'
writeLines(gas7, con = "/tmp/gas8.GeoJSON")
gas8 <- readOGR(dsn = "/tmp/gas8.GeoJSON", layer = "OGRGeoJSON")
proj4string(gas8)
```

This is a messy solution but it's the only one I could find to write
GeoJSON files with a CRS defined for Australian data services' default CRS.










```{r, echo=FALSE}
# library(rgdal)
# library(pryr)
# pryr::show_c_source(readOGR)
# showMethods(readOGR)
# ??readOGR
```


