---
title: "ggmap"
author: "Robin Lovelace"
date: "September 30, 2015"
output: html_document
---

This vignette presents additional mapping options with ggmap.
The content was removed from the main tutorial when the code stopped working.

## Adding base maps to ggplot2 with ggmap

[ggmap](http://journal.r-project.org/archive/2013-1/kahle-wickham.pdf)
is a package that uses the **ggplot2** syntax as a
template to create maps with image tiles taken from map servers
such as Google and
[OpenStreetMap](http://www.openstreetmap.org/):

```{r}
library(ggmap) # install.packages("ggmap") if not installed
```

The `lnd` object loaded previously is in British National Grid but the ggmap
image tiles are in WGS84. We therefore need to use the `lnd84`
object created in the reprojection operation (see Part III).
Load this with `readRDS()`.

```{r}
lnd84 <- readRDS("data/lnd84.Rds") # load previously saved object
```

The first job is to calculate the bounding box (bb for short) of the
`lnd84` object to identify the geographic extent of the image tiles that we need.

```{r}
bb <- bbox(lnd84)
b <- (bb - rowMeans(bb)) * 1.05 + rowMeans(bb)
# scale longitude and latitude (increase bb by 5% for plot)
# replace 1.05 with 1.xx for an xx% increase in the plot size
```

This is then fed into the `get_map` function as the location parameter. The syntax below contains 2 functions. **ggmap** is required to produce the plot and provides the base map data.

```{r, eval=FALSE, message=FALSE}
lnd_b1 <- ggmap(get_map(location = b)) # create basemap for london
```

```{r, echo=FALSE}
# save(lnd_b1, file = "data/lnd_b1.RData")
load("data/lnd_b1.RData") # load saved map
```

In much the same way as we did above we can then layer the plot with different geoms.

First fortify the `lnd84` object and then merge with the required attribute
data (we already did this step to create the `lnd_f` object).

```{r, message=FALSE}
lnd_wgs84_f <- fortify(lnd84, region = "ons_label")
lnd_wgs84_f <- left_join(lnd_wgs84_f, lnd84@data,
  by = c("id" = "ons_label"))
```

We can now overlay this on our base map.

```{r, eval=FALSE}
lnd_b1 +
  geom_polygon(data = lnd_wgs84_f,
  aes(x = long, y = lat, group = group, fill = Partic_Per),
    alpha = 0.5)
```

The code above contains many parameters. Use the **ggplot2** help pages
(e.g. `?geom_polygon`) to find out what they are.
The resulting map could be improved with a simpler basemap in black and white.
Stamen provide the tiles we need and they can be brought into the
plot with the `get_map` function:^[Note that a wide range of customised
on-line basemaps can be used, by modifying 'cloudmade' map styles or
customised Google Maps - see [r-bloggers.com/creating-styled-google-maps-in-ggmap](http://www.r-bloggers.com/creating-styled-google-maps-in-ggmap/) for details on
the latter.]

```{r, eval=FALSE}
# download basemap (use load("data/lnd_b2.RData") if you have no internet)
lnd_b2 <- ggmap(get_map(location = b))
```

```{r, echo=FALSE}
# save(lnd_b2, file = "data/lnd_b2.RData")
load("data/lnd_b2.RData") # load saved map
```

We can then produce the plot as before:

```{r fig.cap="Basemap 2"}
library(mapproj) # mapproj library needed - install.packages("mapproj")

lnd_b2 +
  geom_polygon(data = lnd_wgs84_f, 
    aes(x = long, y = lat, group = group, fill = Partic_Per),
    alpha = 0.5)
```

Finally, to increase the detail of the base map, we can use `get_map`'s `zoom` argument
(result not shown)

```{r, eval=FALSE}
# download basemap (try load("data/lnd_b3.RData") if you lack internet)
lnd_b3 <- ggmap(get_map(location = b, source = "stamen",
  maptype = "toner", crop = TRUE, zoom = 11))
```

```{r, eval=F, echo=FALSE}
# save(lnd_b3, file = "data/lnd_b3.RData")
# load("data/lnd_b3.RData") # load saved map
```

```{r fig.cap="Basemap 3", eval=FALSE}
lnd_b3 +
  geom_polygon(data = lnd_wgs84_f,
    aes(x = long, y = lat, group = group, fill = Partic_Per),
    alpha = 0.5)
```