# Clipping and aggregating with gIntersects

```{r}
library(sp)
library(rgeos)
```


This short piece demonstrates how clipping spatial data can be done using the 
rgeos package in R. A more common method is to use the sp::over function, 
as detailed in another, larger tutorial [see here](https://github.com/Robinlovelace/Creating-maps-in-R/blob/master/intro-spatial-rl.pdf?raw=true).

The test datasets are contained in the data folder of a 
repository called Creating-maps-in-R. These, and the rgeos library
needed for the function can be loaded with the 
following commands:

```{r}
load("data/lnd.RData")
load("data/stations.RData")
```

Here, we use `gIntersects`,
although we could equally use 
`gContains`, `gWithin` and other `g...` functions. 
The power of these commands can be seen by accessing the 
rgeos help pages, e.g. `?gOverlaps`.
`gIntersects` will output information for each point, telling us which 
polygon it interacts with (i.e. the polygon it is in).

Let's take a look at how the function works on our test data:

```{r Identifying and plotting individual stations, results='hide'}
int <- gIntersects(stations, lnd, byid = T) # find which stations intersect 
class(int) # it's outputed a matrix
dim(int) # with 33 rows (one for each zone) and 2532 cols (the points)
summary(int[,c(200,500)]) # not the output of this
plot(lnd)
points(stations[200,], col = "red") # note point id 200 is outside the zones
points(stations[500,], col = "green") # note point 500 is inside
which(int[,500] == T) # this tells us that point 500 intersects with zone 32
points(coordinates(lnd[32,]), col = "black") # test the previous statement
```

In the above code, only the first line actually 'does' anything
in our workspace, by creating the object `int`. The proceeding 
lines are dedicated to exploring this object and what it means. 
Note that it is a matrix with columns corresponding to the points and 
rows corresponding to boroughs. The borough in which a particular 
point can be extracted from `int` as we shall see below.
For the purposes of clipping, we are only interested in whether
the point intersects with _any_ of the boroughs. This is where the 
function `apply`, which is unique to R, comes into play:

```{r Clipped points (within London boroughs)}
clipped <- apply(int == F, MARGIN = 2, all)
plot(stations[which(clipped),]) # shows all stations we DO NOT want
stations.cl <- stations[which(!clipped),] # use ! to select the invers
points(stations.cl, col = "green") # check that it's worked
stations <- stations.cl; rm(stations.cl) # tidy up: we're only interested in clipped ones
```

The first line instructs R to look at each column (`MARGIN = 2`, we would use
`MARGIN = 1` for row-by-row analysis) and report back whether `all` of the values are
false. This creates the inverse selection that we want, hence the use of `!` to invert it.
We test that the function works on a new object (often a good idea, to avoid overwriting 
useful data) with plots and, once content that the clip has worked, save the sample of 
points to our main `stations` object and remove the now duplicated `stations.cl` object.



## Optional task: aggregation with gIntersects

```{r, echo=FALSE}
# This should be a separate vignette/rpubs doc
```

As with clipping, we can also do spatial aggregation with
the **rgeos** package. In some ways, this method makes explicit
the steps taken in `aggregate` 'under the hood'.
The code is quite involved and intimidating, so feel free to
skip this stage. Working through and thinking about it this alternative method may, however,
pay dividends if you intend to perform more sophisticated spatial analysis in R.

```{r}
int <- gIntersects(stations, lnd, byid = TRUE) # re-run the intersection query
head(apply(int, MARGIN = 2, FUN = which))
b.indexes <- which(int, arr.ind = TRUE) # indexes that intersect
summary(b.indexes)
b.names <- lnd$name[b.indexes[, 1]]
b.count <- aggregate(b.indexes ~ b.names, FUN = length)
head(b.count)
```

The above code first extracts the index of the row (borough) for
which the corresponding column is true and then converts this into
names. The final object created, `b.count` contains the number of station
points in each zone. According to this, Barking and Dagenham should contain
12 station points. It is important to check the output makes sense at
every stage with R, so let's check to see if this is indeed the case with
a quick plot:

```{r fig.cap="Transport points in Barking and Dagenham"}
plot(lnd[grepl("Barking", lnd$name),])
points(stations)
```

Now the fun part: count the points in the polygon and report back how many there are!

We have now seen how to load, join and clip data. The second half of this tutorial
is concerned with *visualisation* of the results. For this, we will use
**ggplot2** and begin by looking at how it handles non-spatial data.
