Reading and processing .gpx files
========================================================

## To do
 - Add previous examples of GPX data in R
 - Upload GPX data to OSM

GPS eXchange (.gpx) is the standard format for saving georeferenced (lat/long)
location data taken from Global Positioning System devices.
The file system itself is a type of xml and can be read as plain text.

Hundreds of thousands of .gpx files are now generated each day worldwide,
a massive source of information about where and when processes take place.
Processing, analysing and visualising these .gpx files is thus useful, 
for example for tracing the movements of ships, geolocating photos 
or simply for tracing a trip for fun.

There are many software products designed specifically for handling .gpx
files and most Geographical Information Systems (GIS) applications allow
for .gpx files to be loaded. Being a fan of reproducible research and believer
in the empowering potential of free software, 
I decided to try explore some of my own .gpx data in R, to see what it could do.

This article presents the results of my investigations. It demonstrate
some basic methods for loading, analysing and visualisting .gps data in
the world's favourite free statistical programming language: R.

## Loading the data

The go-to package for loading spatial data in R is
rgdal, using the `readOGR` function. This method has many advantages over
alternative methods, including computational efficiency and the wide range of
formats available: typing `ogrDrivers()[which(ogrDrivers()[,2]),]`
will show you all the wierd and wonderful formats that can be loaded with
your installation - in my
[OSGeo-Live](https://www.bountysource.com/teams/neovim/fundraiser) install 33 formats are supported.
This is how point and track layers can be loaded:


```{r}
library(rgdal)
(layers <- ogrListLayers("data/2014-03-22.gpx"))
gp1 <- readOGR("2014-03-170750_20140317_0750.gpx", layer=layers[5])
gt1 <- readOGR("2014-03-170750_20140317_0750.gpx", layer=layers[3])
names(gp1)
gp1@data[1:5,1:6]
plot(gp1)
plot(gt1, add = T, col = "red")
```

Here we have loaded the data as a point and a track layer (`gp1` and `gt1` respectively)
As the plots illustrate, the track contains a dense network in the west of the plot 
and then a more extensive loop extending to the east. We shall further explore
these data and interpret their meaning in the next section. For now, however, 
let's see another way to load the dataset.

```{r, echo=FALSE, eval=FALSE}
The more convoluted way to load the track is remember that gpx files are 
a type of xml, with classes defined for waypoints, tracks, coordinates etc.
The XML package is used to parse xml data, including scraping websites.
```



```{r}
library(plotKML)
p1 <- readGPX("~/Dropbox/georoutes2/2014-03-22.gpx")
names(p1)
head(p1$tracks[[1]][[1]])
head(unlist(gp1$tracks))
t1 <- do.call("rbind", unlist(p1$tracks, recursive=F))
plot(t1$lon, t1$lat)
## Extract data frames
```

## Extracting basic statistics

It is all very well and good loading and making basic plots of
gpx data. But to be able to extract useful insights from
data and understand something of *what it means*, 
we must do some quantitative analysis. For understanding
the core features of almost any dataset, a surprisingly small
number of functions are needed. To find the number of points
in the dataset and the timespan of data collection, 

```{r}
nrow(gp1)
range(as.character(gp1$time))
```

So we have ~2000 track points collected over the duration
of 3 days, from the 17th untill the 20th March 2014.
GPS devices have very accurate timestamp data: they
rely on fraction of a second differences between different
satalit signals to triangulate position. To extract this
time data, we need to convert the timestamp factors
first into a character string, and finally into a date-time class.

```{r}
library(chron)
gp1$ctime <- as.chron(as.character(gp1$time))
plot(gp1$ctime)
```

To better understand the distribtut

```{r}
gp1.18 <- gp1[which(chron::days(gp1$ctime) == 18),]
plot(gp1.18)
gp1.n18 <- gp1[-which(chron::days(gp1$ctime) == 18),]
```



## A heat-map of activity

A heat map can demonstrate where location is most concentrated,
even in cases where the points or lines are so densly packed that
overlap. A basic heat map of the area can be created very rapidly 
using `spatstat

```{r}
library(spatstat)
library(maptools)
gp1.ppp <- as(SpatialPoints(gp1), "ppp")
plot(density(gp1.ppp))
```

As might have been expected, the results of this preliminary heat map plot are
rather dissapointing: the zone of high density to the west totally dominates,
while the loop is barely visible. The density increases slightly where
the two sides of the loop meet just to east of centre, indicating that
this line is overlapped. The first way to deal with this is to plot the
heatmap with a lower bandwidth, reducing the area over which points are
spatially aggregated.

```{r}
plot(density(gp1.ppp, adjust=0.2))
```

Still it is difficult to extract useful information from this visualisation
because we are essentially dealing processes on two different spatial scales.
The large loop out to the east is many times the area of the high density
blob to the west. So, using the information gleaned in the previous section
we take a subset of the data for future heatmaps to see what is going on in that
western blob.

gp1.

## Looking at elevation

Recalling the contents of the `@data` slot of the data, let's take a quick
look at elevation, which can be useful for various applications.

```{r}
plot( gp1$track_seg_point_id, gp1$ele )
```



