---
title: "Multipart polygons R"
author: "Robin Lovelace"
date: "June 2, 2015"
output: html_document
---

Based on this:
https://stackoverflow.com/questions/30583048/convert-features-of-a-multifeature-geojson-into-r-spatial-objects/30593077#30593077

Normally you can read geojson files into R with trusty `readOGR`, as illustrated [here][1].

However, this fails for multifeature geojsons.

Reproducible example:

```{r}
downloader::download("https://github.com/Robinlovelace/Creating-maps-in-R/raw/master/data/test-multifeature.geojson", "test.geojson")
try(test <- rgdal::readOGR("test.geojson", "OGRGeoJSON")) # fails with:
```

The error message is clear-enough and indicates a solution: split the features.
Aside from doing this with regex, I don't know how, however.

Any ideas very much welcome.

The amazing thing: [GitHub displays the data natively on the browser][2], whereas R cannot even (seemingly) read it in!

Alternative way to a solution:

```{r}
test <- geojsonio::geojson_read("test.geojson")
```

## Solution from Spacedman

```{r}
mess <- 'ogr2ogr -where "OGR_GEOMETRY=\'LINESTRING\'" -f "GeoJSON" lines.geojson  test.geojson'
system(mess)
testl <- rgdal::readOGR("lines.geojson")
```


  [1]: https://github.com/Robinlovelace/Creating-maps-in-R/blob/master/vignettes/geoJSON.Rmd
  [2]: https://github.com/Robinlovelace/Creating-maps-in-R/blob/master/data/test-multifeature.geojson
