---
output: github_document
---

I had to remind myself what `gsection()` does despite packaging it up into **stplanr** (the orginal code was written by Barry Rowlingson). It is used primarily in my work as a helper function for `overline()` but I decided to export it in case it's of use/interest to others. Great to see it is!

The function does not return data for a reason: individual segments have different numbers of overlapping routes.

However, it is useful to be able to query data from which the segments come, so let's work through some code, building on your reproducible example, to see what's going on:

```{r}
library(stplanr)
length(routes_fast) # too many to visualise segments
r = routes_fast[3:4,] # take 2 lines to see what's going on
s = gsection(r) # split into overlapping sections
class(r) # has data, as you say
class(s) # does not have data!
length(r) # 2 lines, as expected
length(s) # 3 segments with same number of overlaps
```

As you can see from the output of the above code chunk, there are more segments than there are routes. So surely each segment can be allocated it's own route? No.

This is illustrated below. The 3rd line from the resulting segments `s` (coloured grey) is the result of the overlap between both lines in `r`. So what data values would you expect it to have?

```{r, warning=FALSE}
library(tmap) # for awesome plotting abilities
qtm(routes_fast[3:4,], line.lwd = 20, line.alpha = 0.3) +
  qtm(routes_fast[3,], line.lwd = 5) +
  qtm(s[1,], line.col = "white") +
  qtm(s[2,], line.col = "black") +
  qtm(s[3,], line.col = "grey", line.lwd = 2)
```

There are different ways to answer this question. The default way in `sp::over()` is to take the first overlap. But this is not what we want as `over()` returns a match even if the lines touch but do not have any shared distance (take a look inside the results to see what I mean):

```{r}
result_data = over(x = s, y = r)
result_data
result_list = over(x = s, y = r, returnList = T)
```

`result_data` returns the first matching row from data in lines touching each segment - in this case that's simply `routes_fast@data[3,]` repeated 3 times, not very useful!

Assuming you're happy with *the first* match of lines that actually share lengths, you could use the (undocumented) `minDimension` argument of `over()`, described in `vignette("over")`:

```{r}
over(x = s, y = r, minDimension = 1)
```

I think adding an argument `return_data` to the function would useful and plan to do so before the next release of **stplanr**. It should probably say something about how many overlapping lines as an option output.

Many thanks for your instigating these investigations in any case: very useful.




