---
title: 'Point Pattern analysis and spatial interpolation with R'
output: github_document
bibliography: references.bib
---

```{r, echo=FALSE}
knitr::opts_knit$set(root.dir = "..")
```

This tutorial teaches the basics of point pattern analysis in R.
It is influenced by the chapter on Spatial Point Pattern Analysis in *Applied Spatial Data Analysis with R* [@bivand_applied_2013] and an [online tutorial](http://rspatial.org/analysis/rst/8-pointpat.html) on Point Pattern Analyis by Robert Hijmans.

```{r}
library(sp)
```

The input datasets are on 'Boris Bikes' docking stations around London, and can be read-in and visualised with the following commands:

```{r cycle-hire1}
cycle_hire = rgdal::readOGR("data/cycle_hire.geojson")
cycle_hire_osm = rgdal::readOGR("data/cycle_hire-osm.geojson")
plot(cycle_hire)
points(cycle_hire_osm, col = "red")
```

It is immediately clear that these two datasets are closely related (they have a high degree of spatial correlation) and have a distinctive pattern.
But how to describe that quantitively, and extrapolate the values from the plotted location to other areas?

It is the purpose of this tutorial to provide the knowhow to answer such questions.


