---
title: "Spatio-temporal data with R"
author: "Robin Lovelace"
date: "May 27, 2016"
output: pdf_document
bibliography: ../creating-maps.bib
---

```{r, echo=FALSE}
knitr::opts_knit$set(root.dir = "../")
```


The most mature package for working with spatio-temporal data in R is probably **spacetime**, which was developed by Edzer Pebesmer, who also developed **sp*:

```{r}
library(spacetime)
```

There are fantastic tutorials on spatio-temporal data in this package, as can be revealed by the `vignettes()` command. Take a browse of these and open

```{r, eval=FALSE}
vignette(package = "spacetime")
vignette("jss816")
```

## The structure of spatial-temporal data

Regardless of software, spatio-temporal data is usually represented in one of three ways [pebesma_spacetime:_2012]:

- Time-wide, where each column is a date.
- Space-wide, where each column is a place
- Long, where each place is repeated for each time

## Example with air quality data

To understand spatio-temporal data in **spacetime**, let's look at an air quality dataset:

```{r}
data(package = "spacetime")
```

```{r}
library(maptools)
fname = system.file("shapes/sids.shp", package="maptools")[1]
nc = readShapePoly(fname,
                   proj4string = CRS("+proj=longlat +datum=NAD27 +ellps=clrk66"))
time = as.POSIXct(c("1974-07-01", "1979-07-01"), tz = "GMT")
endTime = as.POSIXct(c("1978-06-30", "1984-06-30"), tz = "GMT")
data = data.frame(
  BIR = c(nc$BIR74, nc$BIR79),
  NWBIR = c(nc$NWBIR74, nc$NWBIR79),
  SID = c(nc$SID74, nc$SID79))
nct = STFDF(sp = as(nc, "SpatialPolygons"), time, data, endTime)
class(nct)
```

We have created a spacio-temporal datset. Let's look at its componenets:

```{r}
plot(nct@sp) # the spatial component
head(nct@data) # the data with the first 100 rows being for time period 1, etc
nct@time # the time snapshots in there
```

**spacetime** has native methods for natively plotting spatio-temporal data:

```{r}
stplot(nct)
```

### Subsetting spacetime objects

Remember that spatial data can be subsetted using the square brackets:

```
[object_element, column_name]
```

Spatio-temporal data has an extra dimension, so we add another comma:

```
[, time]
```

```{r}
n74 = nct[, "1974-07-01"]
class(n74)
spplot(n74)
```

```{r}
stplot(nct[1:2,]) # select by area
```

## Example with Irish Windspeed data

Let's load the data:

```{r}
library(gstat)
data("wind")
wind.loc$y = as.numeric(char2dms(as.character(wind.loc[["Latitude"]])))
wind.loc$x = as.numeric(char2dms(as.character(wind.loc[["Longitude"]])))
coordinates(wind.loc) = ~x+y
proj4string(wind.loc) = "+proj=longlat +datum=WGS84"
library(tmap)
qtm(wind.loc)
```

Before this is can be a **spacetime** class, we need to update the time variable and perform some more data manipulation:

```{r}
wind$time = ISOdate(wind$year+1900, wind$month, wind$day)
wind$jday = as.numeric(format(wind$time, '%j'))
wind.loc = wind.loc[match(names(wind[4:15]), wind.loc$Code),]
stations = 4:15
windsqrt = sqrt(0.5148 * as.matrix(wind[stations])) # knots -> m/s
Jday = 1:366
windsqrt = windsqrt - mean(windsqrt)
daymeans = sapply(split(windsqrt, wind$jday), mean)
meanwind = lowess(daymeans ~ Jday, f = 0.1)$y[wind$jday]
velocities = apply(windsqrt, 2, function(x) { x - meanwind })
pts = coordinates(wind.loc[match(names(wind[4:15]), wind.loc$Code),])
rownames(pts) = wind.loc$Station
pts = SpatialPoints(pts, CRS("+proj=longlat +datum=WGS84 +ellps=WGS84"))
library(rgdal)
utm29 = CRS("+proj=utm +zone=29 +datum=WGS84 +ellps=WGS84")
pts = spTransform(pts, utm29)
wind.data = stConstruct(velocities, space = list(values = 1:ncol(velocities)),
                        time = wind$time, SpatialObj = pts, interval = TRUE)
class(wind.data)
```



## Exercise

Work through the contents of @pebesma_spacetime:_2012 and generate maps of shifts in wind speed in Ireland.

## References


