---
title: "**stplanr**: A package for transport planning"
author: "Robin Lovelace"
date: "`r Sys.Date()`"
output:
  ioslides_presentation:
    transition: slower
logo: ../figure/its-logo-square.png
---

```{r, include=FALSE}
library(stplanr)
library(sp) # needed for geographical objects
library(leaflet) # needed for plotting data
```

## Introduction

Common computational tasks in transportation planning:

- Access to and processing of data on transport infrastructure and behaviour
- Analysis and visualisation of origin-destination flow data
- Allocation of flows to the transport (road) network
- Development of models to estimate transport behaviour
- The calculation of 'catchment areas' affected by transport infrastructure

## Transparency in transport planning 

Transport planning is notoriously reliant on 'black boxes' and
the same applies to scientific research into transport systems
[@Waddell2002].

**stplanr** seeks to address these issues. After the package
has been installed (see the package's
[README](https://github.com/Robinlovelace/stplanr) for details),
it can be loaded in with `library()`:

```{r, eval=FALSE}
install.packages("stplanr")
library(stplanr)
```

## Accessing and converting data

```{r, echo=FALSE}
d = read.csv(text = "
cents,            SpatialPointsDataFrame of home locations for flow analysis.
destination_zones,example destinations data
destinations,     example destinations data
flow,             data frame of commuter flows
flow_dests,       data frame of invented commuter flows with destinations in a different layer than the origins
flowlines,        SpatialLinesDataFrame of commuter flows
routes_fast,      SpatialLinesDataFrame of commuter flows on the travel network
routes_slow,      SpatialLinesDataFrame of commuter flows on the travel network
zones,            SpatialPolygonsDataFrame of home locations for flow analysis.
           ")
knitr::kable(d)
```
## 'Flow' or OD data

```{r}
data("flow", package = "stplanr")
head(flow[c(1:3, 12)])
```

## Centroids data

```{r}
data("cents", package = "stplanr")
as.data.frame(cents[1:3,-c(3,4)])
```

## Creating 'desire lines'

```{r}
l <- od2line(flow = flow, zones = cents)
# remove lines with no length
l <- l[!l$Area.of.residence == l$Area.of.workplace,]
plot(l, lwd = l$All / 10)
```

