---
title: "Testing rnaturalearth"
author: "Robin Lovelace"
date: "December 15, 2015"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

[rnaturalearth](https://github.com/AndySouth/rnaturalearth) is an R package for accessing a sample of the world's geographic data. It allows quick access to the data stored at [naturalearthdata.com](http://www.naturalearthdata.com/).

It's not yet on CRAN, so you need to install the devtools package to install it easily from github, via:

```{r, eval=FALSE}
install.packages("devtools") # install devtools if not already on your system
devtools::install_github("andysouth/rnaturalearth", build_vignettes=TRUE)
```

`vignette('rnaturalearth')` provides an introduction to its functionality, outlined below.

The functions provided to access data all begin with `ne_`, making it intuitive to use.
Some examples are provided below:

```{r, fig.keep='none'}
library(rnaturalearth) # load the package
greece <- ne_countries(country = "Greece")
library(sp) # load the sp library for plotting
plot(greece)
```

The data that is pulled in varies depending on various arguments:

```{r, fig.keep='none'}
greece2 <- ne_countries(scale = 50, country = "greece")
plot(greece2)
```

This second plot shows that the 50m resolution data is much more detailed and recommended for plotting individual countries. Even higher resolution is available with the 10m dataset.

```{r, fig.keep='none'}
greece3 <- ne_countries(scale = 10, country = "greece")
plot(greece3)
```

These plots are compared in the following plot:

```{r}
par(mfrow = c(1, 1))
plot(greece)
plot(greece2, border = "red", add = T, lwd = 3)
plot(greece3, border = "blue", add = T)
```

Sub-country data is provided by `ne_states`:

```{r}
ukstates <- ne_states(country = "United Kingdom")
plot(ukstates)
```

Additional data can be downloaded with `ne_download`:

```{r}
airports <- ne_download(scale = 10, type = "airports")
roads <- ne_download(scale = 10, type = "railroads", category = "cultural")
railroads <- ne_download(scale = 10, type = "railroads", category = "cultural")
```







