---
title: "tidyr and reshape2"
author: "Robin Lovelace"
date: "11/10/2014"
output: html_document
---

```{r}
library(reshape2)
library(tidyr)
```

The dataset we'll be using to explore the similarities between
**reshape2** and **tidyr** is taken from the excellent book
*R in Action* @kabacoff2011r:

```{r}
df <- read.table(textConnection("1 1 5 6
1 2 3 5
2 1 6 1
2 2 2 4"), sep = " ")
names(df) <- c("id", "time", "X1", "X2")
```

## Reshaping data with reshape2

To prevent `reshape` from using all column headings as variable names,
id variables must be specified. In this case, we want 'time' and 'id'
to act as the id variables, as each is a something that varies.
The values stay the same, but are placed in a single column with a new
column 'variable' replicating the original column names.

```{r}
dfm1 <- melt(df, id.vars = "id")
head(dfm1)
dfm2 <- melt(df, id.vars = c("id", "time"))
dfm2
```

## Tidying the data with tidyr

**tidyr** performs exactly the same reshaping tasks as **reshape2**,
but contains less functionality for aggregating data.
(Data aggregation functions are reserved for **dplyr**.)
The syntax is different: being more concise means the `gather`
function (analogous to `melt` in **reshape2**) can be tricky to learn.
Let's start by replicating the analysis undertaken above.

```{r}
dfg1 <- gather(df, variable, value, -id)
head(dfg1)
dfg2 <- gather(df, variable, value, -id, -time)
```

The `gather` examples may seem strange at first because the syntax
is so different. `gather` requires 4 and 5 arguments to reproduce `dfm1` and `dfm2`
respectively, wheras `melt` requires only two. The extra arguments
are needed because `gather` forces you to specify column names
for the output data frame, rather than the default 'variable' and 'value'
used by `melt`. Note that without new variable names specified, `gather` fails:

```{r}
gather(df, -id) # fail as no column names are specified
gather(df, variable, -id) # fail as only one column name is provided
```

It is also strange that column names in the original dataset
can only be referred to using the minus sign: when I first discovered
`gather`, I kept trying to refer to existing columns, which of course fails.

```{r}
gather(df, time, id)
```

From the above we can deduce the following rule for using gather:

    At least one id variable from the original dataset must be specified with a minus sign and two column names for the output data frame must be provided.
    
## Casting and spreading molten and tidy data

To cast data in reshape2 `dcast` is used:

```{r}
dcast(dfm1, formula = )
```

```{r}
spread(df, X1, X2)
spread(dfg1, id, variable)
```








