---
title: "Intermediate and advanced R for spatial data"
author: "Robin Lovelace"
date: "March 23, 2016"
output: word_document
---

## Description

R is one of the fastest growing languages for data analysis and is increasingly
important in business, government and academia. Less well-known is that R has
a diverse suite of geospatial tools and can function as a fully integrated
Geographical Information System (GIS). Time-series maps, geographically
weighted regression, animations and even online maps become possible when
you understand R's unique language and approach to spatial objects.

However, as with any language, it is important to gain a strong understanding
of the underlying syntax and structure before moving on to complex uses. 
This course therefore focuses on the foundations: how R can be used to load,
manipulate, process, transform and visualise spatial data.

Prior experience with R is highly recommended;
participants will also benefit from experience with GIS.

In the first day we will cover general concepts in efficient R programming and the basics of spatial data classes in R. Building on this strong foundation the second day will cover advanced topics including raster data and the creation of interactive web-based maps using R.

A common thread throughout the course will be plotting geographic data to boost understanding.


```{r, echo=FALSE}
# Attendees with a clearly
# defined problem that R's spatial capabilities can solve will find the course
# especially beneficial due to its applied and pragmatic nature.
```

## Learning outcomes

By the end of the course participants will:

- Be more efficient general R programmers
- Understand the structure of spatial data in R
- Be able to query, subset and analyse spatial objects
- Have a working knowledge of fundamental GIS functions such as changing projections
- Be proficient in the use of R to create maps using add-on packages such as tmap
- Have some experience with advanced functionality such as raster data and interactive maps

## Further information

Cost: TBA for UK registered postgraduate students; TBA for academics/staff at Institutes of Higher Education, public sector staff and staff at registered charity organisations. TBA others

Fee includes lunch and refreshments.

## Agenda

### Day 1: Intermediate R and the basics of R as a GIS

- Registration and refreshments (9:00 - 9:30)

- An introduction to RStudio (9:30 - 10:30) - RL

- Objects, functions and concepts for efficient R programming (10:30 - 11:30) - LC
    - 15 minutes intro, 45 minutes hands on

- Data manipulation and plotting paradigms (11:30 - 12:30) - RL
    - 20 minutes lecture, 40 minutes hands on
    - 

**LUNCH: 12:30 - 13:30**

- Spatial data in R, sp classes and projections (13:30 - 14:30) - LC
    - SpatialPoints
    - SpatialPointsDataFrame
    - Using `spTransform()`

- Loading, plotting and interrogating spatial data, including shapefiles, xy and spatial queries (14:30 - 15:30) - RL
    - Importance of geographic vs projected CRSs

**15:30 - 15:45 Refreshments**

- Manipulating spatial objects with a focus on rgeos (15:45 - 16:00) - LC
    - `gDifference()` and friends

- Making maps with **tmap** (16:00 onwards) - RL

### Day 2: Advanced R for spatial data analysis

- Registration and refreshments (9:00 - 9:30)

- Recap on R as a GIS: a worked example (9:30 - 10:30) - RL
    - Sea Level Rise example 

- Introduction raster data with R (10:30 - 11:30) - LC
    - From first principles

- Raster/vector operations with R - class conversions and aggregation with raster (11:30 - 12:30) - LC

**LUNCH: 12:30 - 13:30**

- Spatio-temporal data with **spacetime** (13:30 - 14:30) - RL

- Points pattern analysis - points to surfaces by IDW, Kriging (14:30 - 15:30) - LC

```{r, echo=FALSE}
# - Transport data with **stplanr** (14:30 - 15:30)
```

**15:30 - 15:45 Refreshments**

- Advanced graphics: online mapping with **mapview** and leaflet and static maps with **tmap** (15:45 - 16:00) - RL

- Taking it further, applications and Q & A (16:00 onwards)

### Prerequisites

Working knowledge of R and RStudio is assumed.

Background reading “A (very) short introduction to R”, Paul Torfs & Claudia Brauer
http://cran.r-project.org/doc/contrib/Torfs+Brauer-Short-R-Intro.pdf

### Course home-page

For more information on the course, please see here: https://github.com/Robinlovelace/Creating-maps-in-R