---
title: "Teaching R as a command line GIS"
author: "Robin Lovelace, `@robinlovelace`"
date: "16/07/2014"
output:
  beamer_presentation:
    highlight: pygments
    incremental: yes
---

## A presentation of 2 halves

```{r, echo=FALSE}
library(png)
library(grid)
#     logo: ~/Dropbox/Public/img/Leeds-logo.jpg
```

```{r Halves, fig.height=2, echo=FALSE}
# img <- readPNG("figure/pitch.png")
#  grid.raster(img) # this presentation: written in R(md)!
```

- Cutting edge research is nothing without communication
- Teaching and outreach benefit from innovation and application

## The two sides of knowledge creation

```{r Sides, fig.height=2.7, echo=FALSE}
img <- readPNG("~/Dropbox/Public/img/ivory-tower.png")
 grid.raster(img) # this presentation: written in R(md)!
```

- Message of this presentation: don't get stuck in an isolated ivory tower or
comfortable teaching bubble!

```{r, echo=FALSE}
# It's a chicken and egg situation: which comes first?
```

## The structure of this presentation

- Inside the Teaching Bubble
    + Problems
    + Solutions
    + Lessons learned   

- Stuck (up) in the Ivory Tower
    + Problems
    + Solutions
    + Lessons learned   

- **Crossing the teaching/learning divide**

```{r, echo=FALSE}
#    + Input data
#    + A spatial interaction model
#    + Extensions

# - Conclusion
```

## A bit of context

Based largely on "visualising spatial data" tutorial, hosted on
CRAN (contributed docs) and github: [**github.com/Robinlovelace/Creating-maps-in-R**](https://github.com/Robinlovelace/Creating-maps-in-R)

Courses delivered through [TALISMAN](http://geotalisman.org/), a
project based at the University of Leeds and UCL. It is 
funded by the ESRC's National Centre for Research Methods ([NCRM](http://www.ncrm.ac.uk/))

## R courses taught

- 7 R courses, 7 months
- 5 introductory, 2 'advanced'
- Over 150 attendees from many different countries

**We improved the content and delivery over time**

**Iteration, feedback and communication are important**

## Human issues

- Very different abilities `->` different learning rates
- Some people 'getting left behind'
- 'Never touched a command line' `->` frustration
- Not understanding method/applications

## Solutions: workshop design

- Extensions for more advanced users
- Reproduce examples in QGIS to make familiar
- Interactive worked examples
- Implement consistent [naming convention](http://www.r-bloggers.com/consistent-naming-conventions-in-r/)

## Technical issues

- Projectors failing
- R or RStudio not installed
- Packages breaking
- No internet
- GitHub `merge`: confusion and chaos

## Solutions to technical issues

- Start VERY slowly
- Right balance of content between:
    + lecturing (boring but necessary)
    + worked examples (risk loosing people but fun and interactive)
    + working on your own (can do at home but encourages good questions)
- Secretly save data that needs to be downloaded from the internet

## Example of 'hidden code' in the tutorial

Tutorial was written in **RMarkdown**, which runs
R code 'live' in the document. Ensures reproducibility
but can create problems.

```{r, message=FALSE, eval=FALSE}
# shown, not run: ```{r, message=FALSE, eval=FALSE}
# create basemap for london
lnd_b1 <- ggmap(get_map(location = b)) 

# not shown but run: ```{r, echo=FALSE}
load("data/lnd_b1.RData") # load saved map
```


## Lessons learned

- The importance of demonstrators
- Repition
- Details are important
- Focus on software more than concepts
- Simplicity


## Part II: Research and what R can do

- World class data manipulation and processing
- Wide range of GIS functions
- **Excels and spatial statistics**
- **Modelling**: e.g. Bayesian model comparison

**Next: some examples**

## Custom data processing:

It can take data in a wide range of formats.
E.g. MySQL database dump gives you this file `ps`:

    LINESTRING(-1.81 52.55,-1.81 52.55, … )  
    LINESTRING(-1.34 52.43,-1.41 52.45, … )  

Solution:

```{r, eval=FALSE}
ps <- as.list(ps) # make a list
for(i in 1:length(ps)){
  ps[[i]] <- gsub("LINESTRING\\(", "", ps[[i]])
  ps[[i]] <- gsub("\\)", "", ps[[i]])
  ps[[i]] <- gsub(" ", ",", ps[[i]])
  ps[[i]] <- matrix(ps[[i]], ncol=2, byrow=T)
  ps[[i]] <- Line(ps[[i]])
}
```

## Themes that cut across the divide

- Reproducibility a great benefit in teaching and research
- Good research can lead to good teaching and vice versa
- The balance between spontaneity and stability
- Describe not only *what* you did, but *how* you did it
    + Details are important - e.g. autocompletion in RStudio

## Transcending the teaching/research divide

- Teaching and research needn't compete
- Good research can provide excellent teaching materials
- Innovative teaching can lead to new research ideas and people power
- Teaching can link people up
- Open source software joins people from different place together

## Conclusions

- **R is Free Open Source Software**
- It is great with geo-spatial data **analysis** and **modelling**
- It's **FOSS4GDAM**! (FOSS4G for short)

- **It deserves more attention from OSGeo and FOSS4G community**
    + E.g. manageR plugin for
    [**integration in QGIS**](http://www.maths.lancs.ac.uk/~rowlings/Teaching/UseR2012/introduction2Talk.html)
    + E.g. R integration with QGIS with [**spqr**](http://www.maths.lancs.ac.uk/~rowlings/Qgis/Plugins/Documentation/spqr/)

- Hopefully this talk will help integration with R and FOSS4G


