---
title: "Introduction to Spatial Data in R"
author: "Robin Lovelace"
output: ioslides_presentation
---

```{r setup, include=FALSE}
library(sp)
library(raster)
```

# Spatial data in R, sp classes and projections 

## The slots of spatial data

```{r}
load("../data/lnd.RData")
slotNames(lnd)
lnd@bbox
```

## Seeing the geometry

```{r}
lnd@polygons[[1]]@Polygons[[1]]@coords[1:5,]
plot(lnd@polygons[[1]]@Polygons[[1]]@coords)
```

## Seeing the data

```{r}
head(lnd@data[1:2,1:3])
```

## Subsetting spatial objects

```{r}
lnd5 = lnd[5,]
plot(lnd5)
```

## What just happened?

A new object was created, a subset of `lnd`, with the same class.

```{r}
class(lnd)
class(lnd5)
nrow(lnd)
nrow(lnd5)
```

## Subsetting based on attributes

We can select spatial objects based on their attributes. Sometimes it can help to first create a selection object:

```{r}
sel = lnd$Pop_2001 > median(lnd$Pop_2001)
lnd_highpop = lnd[sel,]
plot(lnd_highpop)
```

## Adding together different selections

```{r}
plot(lnd)
plot(lnd_highpop, add = TRUE, col = "red")
```

Note: it would be better to do this with **tmap**.

## Illustration with tmap

```{r}
library(tmap)
lnd$colour = "white"
lnd$colour[sel] = "black"
qtm(lnd, fill = "colour")
```







# Loading, plotting and interrogating spatial data | Pages 8 - 10

## Transformations

```{r}
lnd = raster::shapefile("../data/lnd-stns.shp")
proj4string(lnd)
bbox(lnd)
lnd_osgb = spTransform(lnd, CRSobj = "+init=epsg:27700")
bbox(lnd_osgb)
```

# Manipulating spatial objects | Pages 11 - 17

