Failed cluster attempts
========================================================

Trying to get smooth polygons:

```{r}
lns <- slot(SLDF[6,], "lines")
plot(SLDF[6,])
SLDm <- lapply( SLDF@lines , slot , "Lines" )
plot(SLDm) ## fail
```

Try to convert lines to polys, Bivand style:

```{r}
SLDH <- SLDF[6,]
lns <- slot(SLDH, "lines")
table(sapply(lns, function(x) length(slot(x, "Lines"))))

islands <- sapply(lns, function(x) { # fail
crds <- slot(slot(x, "Lines")[[1]], "coords")
identical(crds[1, ], crds[nrow(crds), ])
})


list_of_Lines <- slot(islands, "lines")
islands_sp <- SpatialPolygons(lapply(list_of_Lines, function(x) {
Polygons(list(Polygon(slot(slot(x, "Lines")[[1]],
"coords"))), ID = slot(x, "ID"))
}), proj4string = proj4string(lnd))
summary(islands_sp)


islands <- sapply(lns, function(x) {
crds <- slot(slot(x, "Lines")[[1]], "coords")
identical(crds[1, ], crds[nrow(crds), ])
})
table(islands)
```

table(sapply(lns, function(x) length(slot(x, "Lines"))))

lclusts <- sapply(lns, function(x) {
crds <- slot(slot(x, "Lines")[[1]], "coords")
identical(crds[1, ], crds[nrow(crds), ])
})
table(lclusts)
lls <- SLDF[6,lclusts]
plot(lls)
plot(lclusts, col="red")

plot(SLDF[6,], add=T)
lbc <- gIntersection(lb, SLDF[6,])
plot(lbc)
```

Let's try it with `clusterset`:
```{r}
example(clusterset)
plot(clusterset(sSp, result="domain", ))
```

Extracting it directly from the density:
```{r}
summary(Dens$v)[,1:3]
Dsub <- Dens[which(Dens$v > 1.675e-07)]
summary(Dsub)
plot(Dsub)
```

Using the lines to clip a polygon of London.

```{r}
# lb <- unionSpatialPolygons(lnd, IDs=1:nrow(lnd) ) # don't seem to work
library(rgeos)
lb <- gUnaryUnion(lnd)
library(raster)
Dr <- raster(Dsg)
proj4string(Dr) <- proj4string(stations)
Drc <- cut(Dr, breaks=8) 
Drp <- rasterToPolygons(Drc, dissolve=T) # create density polygons
spplot(Drp) # check that the clustering worked OK
lclust <- Drp[3,]

plot(lb)
plot(lclust, add = T)
plot(stations[lclust,], add=T, col = "red")
```
