Introduction to Spatial Data and ggplot2
========================================================

Building on the introduction provided by the previous worksheet, the next set of exercises are concerned with specific functions for spatial data and also the use of a package called ggplot2 for data visualisation.

## Spatial Data

R has a huge (and growing) number of spatial data packages. On your own machine these are easily installed with the help of the "ctv" package.

```{r, eval=FALSE}
install.packages("ctv")
library(ctv)
# install.views(“spatial”) # This step will download and install all the spatial packages available in R.
# You will not need to do this step today, so it is commented out
```

You will need to download the practical’s data from here: 

https://www.dropbox.com/sh/0z9a0hrn72poql5/Bx3rgWZ0kN

Save this to a new folder, then in R specify the path of that folder as you working directory.
If your username is "username" and you saved the files into a 
folder called "rmapping" into your Desktop, for example, 
you would type the following:

```{r, eval= F}
setwd("C:/Users/username/Desktop/rmapping/R")
```

If you are working in RStudio, it is worth setting up a project that will automatically 
set your working directory. One of the most important steps in handling spatial data with R 
is the ability to read in shapefiles. There are a number of ways to do this. 
The most simple is `readShapePoly()` in the `maptools` package:

```{r, eval=FALSE}
library(maptools) # load the package
sport <- readShapePoly("london_sport.shp") # read in the shapefile
```

This method works OK, but it is no longer considered best 
practice since it doesn’t load in the spatial referencing information 
etc associated with the shapefile.
A more powerful way to read in geographical data is to use the `rgdal` function 
`readOGR`, which automatically extracts this information.
This is R’s interface to the "Geospatial Abstraction Library (GDAL)"
which is used by other open source GIS packages such as QGIS and enables 
R to handle a broader range of spatial data formats.

```{r}
library(rgdal)
sport <- readOGR(dsn = ".", "london_sport")
```

In the code above dsn stands for "data source name" which is often the folder containing the spatial data – this was pre-specified when you set your working directory – and then the text in " " following the comma is the name of the file required. There is no need to add a file extension. The file contains the borough population and the percentage of the population engaging in sporting activities, and was taken from the file "active-people-survey-participation" available from
[data.london.gov.uk](http://data.london.gov.uk/datastore/package/active-people-survey-participation).
The boundary data is from the OS Opendata Scheme: http://www.ordnancesurvey.co.uk/oswebsite/opendata/ .

All shapefiles have an attribute table. This is loaded with `readOGR` and can be treated in a similar way to a `data.frame`. 

R hides the geometry of spatial data unless you print the object (using the `print()`). 
Let's take a look at the headings of sport, using the following command: `names(sport)`
The data contained in spatial data are kept in a 'slot' that can be accessed using 
the @ symbol:

```{r, eval=FALSE}
sport@data
```

This is useful if you do not wish to work with the spatial components of the data at all times. 

Type `summary(sport)` to get some additional information about the data object. Spatial objects in R contain a variety of additional information:

```
Object of class SpatialPolygonsDataFrame
Coordinates:
       min      max
x 503571.2 561941.1
y 155850.8 200932.5
Is projected: TRUE 
proj4string :
[+proj=tmerc +lat_0=49 +lon_0=-2 +k=0.9996012717 +x_0=400000 +y_0=-100000
+ellps=airy +units=m +no_defs]
```

In the above code `proj4string` represents the coordinate reference system used in the data. 
In this file it has been incorrectly specified so we can change it with the following:

```{r}
proj4string(sport) <- CRS("+init=epsg:27700")
```

You will see you get a warning. This is saying that you are simply changing the coordinate reference system and not reprojecting the data. Epsg:27700 is the code for British National Grid. If we wanted to reproject the data into something like WGS84 for latitude and longitude we would use the following code:

```{r}
sport.wgs84 <- spTransform(sport, CRS("+init=epsg:4326"))
```

The different epsg codes are a bit of hassle to remember but you can find them all here: 
http://spatialreference.org/

## ggplot2

This next section of the practical introduces a slightly different method of creating plots in R using the ggplot2 
package. The package is an implementation of Leland Wilkinson's Grammar of Graphics - 
a general scheme for data visualization that breaks up graphs into semantic components such as scales and layers. 
ggplot2 can serve as a replacement for the base graphics in R (the functions you have been plotting with today) and contains a number of default options that match good visualisation practice.

The maps we produce will not be that meaningful - 
the focus here is on sound visualisation with R and not sound analysis 
(obviously the value of the former diminished in the absence of the latter!)
Whilst the instructions are step by step you are encouraged to deviate from them 
(trying different colours for example) to get a better understanding of what we are doing. 

`ggplot2` is one of the best documented packages in R. 
The full documentation for it can be found online and it is recommended you 
test out the examples on your own machines and play with them:

http://docs.ggplot2.org/current/

here is also a cookbook for R with some nice examples:

http://wiki.stdout.org/rcookbook/Graphs/

Load the packages:

```{r}
library(ggplot2)
```

It is worth noting that the basic `plot()` function requires no 
data preparation but additional effort in colour selection/adding the map key etc. 
`qplot()` and `ggplot()` (from the ggplot2 package) 
require some additional steps to format the spatial data but select 
colours and add keys etc automatically. More on this later.

As a first attempt with ggplot2 we can create a scatter plot with the attribute data in the sport object created above. Type:

```{r}
p <- ggplot(sport@data, aes(Partic_Per, Pop_2001))
```

What you have just done is set up a ggplot object where you say where you want the input data to come from. `sport@data` is actually a data frame contained within the 
wider spatial object `sport` (the `@` enables you to access the attribute table of the 
sport shapefile).  The characters inside the `aes` argument
refer to the parts of that data frame you wish to use (the variables `Partic_Per` and `Pop_2001`).
This has to happen within the brackets of `aes()`, which means, 
roughly speaking 'aesthetics that vary'.  
If you just type p and hit enter you get the error `No layers in plot`. 
This is because you have not told ggplot what you want 
to do with the data. We do this by adding so-called "geoms", 
in this case `geom_point()`.

```{r}
p + geom_point()
```

Within the brackets you can alter the nature of the points. Try something like `p + geom_point(colour = "red", size=2)` and experiment. 

If you want to scale the points by borough population and colour them by sports participation this is also fairly easy by adding another `aes()` argument.

```{r}
p+geom_point(aes(colour=Partic_Per, size=Pop_2001))
```


The real power of ggplot2 lies in its ability to add layers to a plot. In this case we can add text to the plot.

```{r}
p + geom_point(aes(colour = Partic_Per, size = Pop_2001)) + geom_text(size = 2, aes(label = name))
```

This idea of layers (or geoms) is quite different from the standard plot functions in R, but you will find that each of the functions  does a lot of clever stuff to make plotting much easier (see the documentation for a full list). 
 
The following steps will create a map to show the percentage of the population in each London Borough who regularly participate in sports activities. 

To get the shapefiles into a format that can be plotted we have to use the `fortify()` function. Spatial objects in R have a number of slots containing the various items of data (polygon geometry, projection, attribute information) associated with a shapefile. Slots can be thought of as shelves within the data object that contain the different attributes. The "polygons" slot contains the geometry of the polygons in the form of the XY coordinates used to draw the polygon outline. The generic plot function can work out what to do with these, ggplot2 cannot. We therefore need to extract them as a data frame. The fortify function was written specifically for this purpose.

```{r}
sport.f <- fortify(sport, region = "ons_label")
```

This step has lost the attribute information associated with the sport object. We can add it back using the merge function (this performs a data join). To find out how this function works look at 
the output of typing `?merge`.

```{r}
sport.f <- merge(sport.f, sport@data, by.x = "id", by.y = "ons_label")
```

Take a look at the `sport.f` object to see its contents.  You should see a large data frame containing the latitude and longitude (they are actually eastings and northings as the data are in British National Grid format) coordinates alongside the attribute information associated with each London Borough. If you type `print(sport.f)` you will just how many coordinate pairs are required!
To keep the output to a minimum, take a peak at the object just using the `head` command:

```{r}
head(sport.f)
```

It is now straightforward to produce a map using all the built in tools 
(such as setting the breaks in the data) that ggplot2 has to offer. 
`coord_equal()` is the equivalent of asp=T in regular plots with R:

```{r}
Map <- ggplot(sport.f, aes(long, lat, group = group, fill = Partic_Per)) + 
  geom_polygon() + 
  coord_equal() + 
  labs(x = "Easting (m)", y = "Northing (m)", fill = "% Sport Partic.") + 
  ggtitle("London Sports Participation")
```

Now, just typing `Map` should result in your first ggplot-made map of London!
There is a lot going on in the code above, so think about it line by line:
what has each of the elements of code above has been designed to do. 
Also note how the `aes()` components can be combined into one set of brackets 
after `ggplot`, that has relevance for all layers, rather than being
broken into separate parts as we did above. 
The different plot functions still know what to do with these. 
The `group=group` points ggplot to the group column added by 
`fortify()` and it identifies the groups of coordinates that pertain 
to individual polygons (in this case London Boroughs). 

The default colours are really nice but we may wish to produce the map in black and white, 
which should produce a map like that shown below:

```{r}
Map + scale_fill_gradient(low = "white", high = "black")
```

Saving plot images is also easy. You just need to use `ggsave` after each plot, e.g.
`ggsave("my_map.pdf")` will save the map as a pdf, with default settings. For 
a larger map, you could try the following:

```{r, eval=FALSE}
ggsave("my_large_plot.png", scale = 3, dpi = 400)
```


## Adding basemaps to ggplot2 with ggmap

ggmap is a package that uses the ggplot2 syntax as a 
template to create maps with image tiles from the likes of Google and OpenStreetMap:

```{r}
library(ggmap) # you may have to use install.packages to install it first
```

The sport object is in British National Grid but the ggmap 
image tiles are in WGS84. We therefore need to uses the sport.wgs84 
object created in the reprojection operation earlier. 

The first job is to calculate the bounding box of the 
sport.wgs84 object to identify the geographic extent of the image tiles that we need. 

```{r}
b <- bbox(sport.wgs84) 
```

This is then fed into the `get_map` function as the location parameter. The syntax below contains 2 functions. `ggmap` is required to produce the plot and provides the basemap data.

```{r, message=FALSE}
lnd.b1 <- ggmap(get_map(location = b))
```

In much the same way as we did above we can then layer the plot with different geoms. 

First fortify the sport.wgs84 object and then merge with the required attribute
data (we already did this step to create the sport.f object).

```{r}
sport.wgs84.f <- fortify(sport.wgs84, region = "ons_label")
sport.wgs84.f<- merge(sport.wgs84.f, sport.wgs84@data, 
                      by.x = "id", by.y = "ons_label")
```


We can now overlay this on our basemap.

```{r, eval=FALSE}
lnd.b1 + 
  geom_polygon(data = sport.wgs84.f, 
               aes(x = long, y = lat, group = group, fill = Partic_Per), 
               alpha = 0.5)
```

The code above contains a lot of parameters. Use the ggplot2 help pages to find out what they are. 
The resulting map looks okay, but it would be improved with a simpler basemap in black and white. 
A design firm called stamen provide the tiles we need and they can be brought into the 
plot with the `get_map` function:

```{r}
lnd.b2 <- ggmap(get_map(location = b, source = "stamen", maptype = "toner", crop = T))
```

We can then produce the plot as before.

```{r}
lnd.b2 + 
  geom_polygon(data = sport.wgs84.f, 
               aes(x = long, y = lat, group = group, fill = Partic_Per),
               alpha = 0.5)
```

Finally, if we want to increase the detail of the basemap, get_map has a zoom parameter.

```{r}
lnd.b3 <- ggmap(get_map(location = b, source = "stamen", maptype = "toner", crop = T, zoom = 11))

lnd.b3 + 
  geom_polygon(data = sport.wgs84.f, 
               aes(x = long, y = lat, group = group, fill = Partic_Per), 
               alpha = 0.5)
```


## Using ggplot2 for Descriptive Statistics

For this we will use a new dataset:

```{r}
input <- read.csv("ambulance_assault.csv")
```

This contains the number of ambulance callouts to assault incidents (downloadable from the London DataStore) between 2009 and 2011. 

Take a look at the contents of the file:

```{r}
head(input)
```

We can now plot a histogram to show the distribution of values. 

```{r}
p.ass <- ggplot(input, aes(x = assault_09_11))
```

Remember the `ggplot(input, aes(x=assault_09_11))` section means create a generic plot object (called p.ass) from the input object using the `assault_09_11` column as the data for the x axis. To create the histogram you need to tell R that this is what you want to go with

```{r, eval=FALSE}
p.ass + geom_histogram()
```

The resulting message (`stat_bin: binwidth defaulted to range/30...`)
relates to the bins - the breaks between histogram blocks.
If you want the bins (and therefore the bars) to be thinner 
(i.e. representing fewer values) you need to make the bins 
smaller by adjusting the binwidth. Try:

```{r, eval=FALSE}
p.ass + 
  geom_histogram(binwidth = 10) + 
  geom_density(fill = NA, colour = "black") 

```

It is also possible to overlay a density distribution over the top of the histogram. 
For this we need to produce a second plot object with the density distribution as the y variable.

```{r}
p2.ass <- ggplot(input, aes(x = assault_09_11, y = ..density..))

p2.ass + 
  geom_histogram() +
  geom_density(fill = NA, colour = "red")
```

What kind of distribution is this plot showing? You can see that there are
a few wards with very high assault incidences (over 750). 
To find out which ones these are we can select them.

```{r}
input[which(input$assault_09_11>750),]
```


It is perhaps unsurprising that St James's and the West End have the highest counts.
The plot has provided a good impression of the overall distribution,
but what are the characteristics of each distribution within the Boroughs?
Another type of plot that shows the core characteristics of the distribution
is a box and whisker plot. These too can be easily produced in R 
(you can't do them in Excel!). We can create a third plot object 
(note that the assault field is now y and not x):

```{r}
p3.ass <- ggplot(input, aes(x = Bor_Code, y = assault_09_11))
```


and convert it to a boxplot.

```{r}
p3.ass + geom_boxplot()
```

Perhaps this would look a little better flipped round.

```{r}
p3.ass + geom_boxplot() +
  coord_flip()
```

Now each of the borough codes can be easily seen. 
No surprise that the Borough of Westminster (00BK) 
has the two largest outliers. In one line of code you 
have produced an incredibly complex plot rich in information. 
This demonstrates why R is such a useful program for these kinds of statistics. 

If you want an insight into some of the visualisations you can develop with this type of data we can do faceting based on the example of the histogram plot above. 

```{r, message=FALSE}
p.ass + geom_histogram() +
  facet_wrap( ~ Bor_Code)
```

We need to do a little bit of tweaking to make this plot publishable but we want to demonstrate that it is really easy to produce 30+ plots on a single page! Faceting is an extremely powerful way of visualizing multidimensional datasets and is especially good for showing change over time.

## Advanced Task: Facetting for Maps

```{r}
library(reshape2) # this may not be installed. 
# If not install it, or skip the next two steps…

```

Load the data - this shows historic population values between 1801 and 2001 for London, again from the London data store.

```{r}
london.data <- read.csv("census-historic-population-borough.csv")
```

"Melt” the data so that the columns become rows.

```{r}
london.data.melt <- melt(london.data, id = c("Area.Code", "Area.Name"))
```


Only do this step if reshape and melt failed

```{r, eval=F}
london.data.melt<-read.csv("london_data_melt.csv")
```


Merge the population data with the London borough geometry contained within our sport.f object.

```{r}
plot.data <- merge(sport.f, london.data.melt, by.x = "id", by.y = "Area.Code")
```

Reorder this data (ordering is important for plots).

```{r}
plot.data <- plot.data[order(plot.data$order), ]
```

We can now use faceting to produce one map per year (this may take a little while to appear). 

```{r}
ggplot(data = plot.data, aes(x = long, y = lat, fill = value, group = group)) + 
  geom_polygon() + 
  geom_path(colour="grey", lwd=0.1) + 
  coord_equal() +
  facet_wrap(~variable)
``` 

Again there is a lot going on here so explore the documentation to make sure you understand it. 
Try out different colour values as well. 

Add a title and replace the axes names with “easting” and 
‘northing” and save your map as a pdf.
