# R for Spatial Data: Futureproof foundations

- 6 October 2017

This document provides a how for the course on 2017-10-06 in LIDA Leeds.

It is primarily intended to provide a set of links for course participants but can also host code examples and notes from the workshop.


## Course feedback

Please fill-in this form to help us improve the course:
 https://leeds.onlinesurveys.ac.uk/rspatial2017 

## Slides

- See the 'slides' folder in for the text: [github.com/Robinlovelace/geocompr](https://github.com/Robinlovelace/geocompr/blob/master/slides/)
- For compiled slides on Spatial Data and the Tidyverse see [robinlovelace.net/presentations/spatial-tidyverse.html](http://robinlovelace.net/presentations/spatial-tidyverse.html#1)

## Links

- Agenda  and other course details on GitHub at [course-info/intro-info.Rmd](https://github.com/Robinlovelace/Creating-maps-in-R/blob/master/course-info/intro-info.Rmd)
- [Geocomputation with R](http://robinlovelace.net/geocompr/)

- An introduction to geographic data in R
    - Chapter 2 of [Geocomputation with R](http://robinlovelace.net/geocompr/spatial-class.html)
- Geographic data I/O
    - Chapter 5 of [Geocomputation with R](http://robinlovelace.net/geocompr/read-write.html)
- Introduction to visualising spatial data with R
    - [Creating-maps-in-R](https://github.com/Robinlovelace/Creating-maps-in-R) GitHub tutorial
- Point pattern analysis and rasterization
    - [Point Pattern analysis and spatial interpolation with R
](https://github.com/Robinlovelace/Creating-maps-in-R/blob/master/vignettes/point-pattern.md) from the previous tutorial

## Demos

```{r}
1:9
x = 1:9
y = x^2
plot(x, y)
library(spData)
library(sf)
library(tidyverse)

world
data(world)
world_df = as_data_frame(world)
```

Edit a file from the command line:

```{r, eval=FALSE}
file.edit("course-info/course-info.Rmd")
```

sp to sf translation

```{r}
library(sp)
g = SpatialGrid(GridTopology(c(0, 0), c(1, 1), c(3, 3)))
p = as(g, "SpatialPolygons")
library(sf)
p_sf = st_as_sf(p)
```

An intro to raster

```{r}
library(spDataLarge)
library(raster)
raster_filepath = system.file("raster/srtm.tif", package = "spDataLarge")
new_raster = raster(raster_filepath)
new_raster
class(new_raster)
plot(new_raster)
mapview::mapview(new_raster)
extent(new_raster)
library(spData)
st_bbox(world)
```



```{r}
library(sf)
library(raster)
library(tidyverse)
library(spData)
world_unite = world %>%
  tidyr::unite("con_reg", continent:region_un, sep = ":", remove = TRUE)

dplyr::select(world, continent)
```

Link to Craig's stuff:

https://github.com/clmorton/HEV

## Spatial operations

Spatial subsetting.

```{r}
library(sf)
library(tidyverse)
library(spData)

africa = world %>% 
  filter(continent == "Africa")
# %>% 
#   filter(grepl("k", name_long))
plot(africa)

africa = filter(.data = world, continent == "Africa")

africa = world[world$continent == "Africa", ]

plot(africa)
plot(africa$geom)
bb = st_bbox(africa)
set.seed(2017)
lat_random = runif(n = 100, min = bb[1], max = bb[3])
lon_random = runif(n = 100, min = bb[2], max = bb[4])
mat_random = cbind(lat_random, lon_random)
p = st_multipoint(x = mat_random) %>% 
  st_sfc() %>% 
  st_cast(to = "POINT") %>% 
  st_sf()
# without the pipe %>% %>% 
# st_sf(st_cast(x = st_sfc(st_multipoint(x = mat_random)) , to = "POINT"))
plot(africa$geom)
plot(p, add = T)

p_in_africa = p[africa, ] # error
st_crs(p) = st_crs(africa)
p_in_africa = p[africa, ]
plot(p_in_africa, add = T, cex = 3)

# worked question: how to do this in the tidyverse?
# work on chapter 4: spatial operations



center = st_sf(geometry = st_sfc(st_point(c(0, 0)), crs = 4326))
buff = st_buffer(x = center, dist = 20)
plot(buff)


```

## Spatial operations

```{r}
a_poly = st_polygon(list(rbind(c(-1, -1), c(1, -1), c(1, 1), c(-1, -1))))
a = st_sfc(a_poly)

# l_line = st_linestring(x = matrix(c(-1, -1, -0.5, 1), , 2))
# l = st_sfc(l_line)

p_matrix = matrix(c(0.5, 1, -1, 0, 0, 1, 0.5, 1), ncol = 2)
p_multi = st_multipoint(x = p_matrix)
p = st_sf(st_cast(st_sfc(p_multi), "POINT"))

plot(a, border = "red", col = "grey", axes = TRUE)
plot(l, add = TRUE)
plot(p, add = TRUE, lab = 1:4)
text(p_matrix[, 1] + 0.1, p_matrix[, 2] - 0.1, 1:4)

st_intersects(p, a)

st_intersects(p, a, sparse = FALSE)

sel = st_intersects(p, a, sparse = FALSE)[, 1]
p[sel, ]

st_disjoint(p, a, sparse = FALSE)[, 1]

st_within(p, a, sparse = FALSE)

st_touches(p, a, sparse = FALSE)
```


## Getting development versions of pkgs
 
Problem: no access to urban_agglomerations dataset
 
Solution: install latest version of spData
 
```{r}
devtools::install_github("nowosad/spData")
```

## Data I/O

- Online

```{r}
url = file.path("http://www.naturalearthdata.com/http//www.naturalearthdata.com",
                "download/10m/cultural/ne_10m_parks_and_protected_lands.zip")
download.file(url = url,
              destfile = "USA_parks.zip")
unzip(zipfile = "USA_parks.zip")
usa_parks = st_read("ne_10m_parks_and_protected_lands_area.shp")
files_to_remove = list.files(pattern = "ne_10m_parks")
file.remove(files_to_remove)
```

- with osmdata

```{r}
install.packages("osmdata")
library(osmdata)
library(dplyr)
q = opq(bbox = "leeds uk") %>% 
  add_osm_feature(key = "leisure", value = "park")
parks = osmdata_sf(q)
parks_polygon = parks$osm_polygons
plot(parks_polygon$geometry)
mapview::mapview(parks_polygon)
```


 
## Advanced topological relations
 
 Selecting 'rook' and 'queen' relations
Source: 

```{r}
library(sp)
g = SpatialGrid(GridTopology(c(0, 0), c(1, 1), c(3, 3)))
p = as(g, "SpatialPolygons")
library(sf)
library(tidyverse)
p_sf = st_as_sf(p)
st_rook = function(a, b = a) st_relate(a, b, pattern = "F***1****")
st_queen <- function(a, b = a) st_relate(a, b, pattern = "F***T****")
p_queen = p_sf %>%
  mutate(NB_ROOK = st_queen(., b = p_sf[5,])) %>%
  filter(NB_ROOK == 1)
nrow(p_queen)
p_rook = p_sf %>%
  mutate(NB_ROOK = st_rook(., b = p_sf[5,])) %>% 
  filter(NB_ROOK == 1)
nrow(p_rook)
p_minDim2 = p_sf %>%
  filter(st_within(., y = p_sf[5,], sparse = FALSE)) 
nrow(p_minDim2)
```